% The bishop and the presbyter, a question recorded. The afterword
% subsection of Mere Catholicity, published also as its own paper. The
% body is extracted verbatim from confession.tex; edit the book, then
% re-extract, rather than editing bishop-presbyter-body.tex directly.
\documentclass[11pt,letterpaper]{article}
\ifdefined\pdfsuppressptexinfo\pdfsuppressptexinfo=-1 \fi
\ifdefined\pdftrailerid\pdftrailerid{}\fi
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{mathpazo}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{needspace}
\usepackage[colorlinks=true,linkcolor=ink,urlcolor=maroon,citecolor=ink,filecolor=ink]{hyperref}
\urlstyle{same}
\hypersetup{pdftitle={The bishop and the presbyter, a question recorded},pdfauthor={Anonymous}}
\linespread{1.06}
\setlength{\emergencystretch}{1.5em}

\definecolor{ink}{HTML}{2A2521}
\definecolor{heading}{HTML}{8C4A32}
\definecolor{subhead}{HTML}{6E5642}
\definecolor{accent}{HTML}{B0894C}
\definecolor{maroon}{HTML}{7B2E2E}

\newcommand{\unit}[1]{\needspace{3\baselineskip}\par\medskip{\normalfont\bfseries\color{subhead}#1}\par\nobreak\smallskip}

\begin{document}
\color{ink}

\begin{center}
{\LARGE\bfseries\color{heading} The bishop and the presbyter,\par}
\vspace{0.3em}
{\LARGE\bfseries\color{heading} a question recorded\par}
\vspace{0.8em}
{\itshape\color{subhead} From the afterword of Mere Catholicity.\par}
\vspace{0.5em}
{\color{accent}\hrule height 1pt}
\end{center}
\vspace{1em}

\noindent\emph{Many readers arrive here with one plain question. Do the
pastors of the Lutheran and Reformed churches, who have no bishops in the
old succession, have valid orders? The honest answer is that the authors
of this confession do not fully agree, and rather than hide the difference
they record it. One reads the episcopate with St.\ Jerome, as the ruling
presbyterate, so that presbyteral orders bear a real defect rather than
nothing at all. Another reads it with Newman, as an authentic development
received by the whole Body. Both readings are set out below at their full
strength, with the patristic and even papal evidence each rests on, so
that the reader is given the weighing rather than a verdict.}

\medskip

\input{bishop-presbyter-body.tex}

\vspace{2em}
{\small\noindent This work is dedicated to the public domain (Creative
Commons CC0 1.0). No rights reserved. It may be copied, translated,
printed, and distributed freely, by anyone, for any purpose.
\href{https://merecatholicity.com}{merecatholicity.com}\par}

\end{document}
