\xchapter{Epistle to the Ephesians}

\textsc{Shorter.}~\emph{Ignatius, who is also called Theophorus, to the Church which is at Ephesus, in Asia, deservedly most happy, being blessed in the greatness and fulness of God the Father, and predestinated before the beginning}\footnote{Literally, “before the ages.”}\emph{ of time, that it should be always for an enduring and unchangeable glory, being united}\footnote{These words may agree with “glory,” but are better applied to the “Church.”}\emph{ and elected through the true passion by the will of the Father, and Jesus Christ, our God: Abundant happiness through Jesus Christ, and His undefiled grace.}

\textsc{Longer.}~\emph{Ignatius, who is also called Theophorus, to the Church which is at Ephesus, in Asia, deservedly most happy, being blessed in the greatness and fulness of God the Father, and predestinated before the beginning}\footnote{Literally, “before the ages.”}\emph{ of time, that it should be always for an enduring and unchangeable glory, being united}\footnote{These words may agree with “glory,” but are better applied to the “Church.”}\emph{ and elected through the true passion by the will of God the Father, and of our Lord Jesus Christ our Saviour: Abundant happiness through Jesus Christ, and His undefiled joy.}\footnote{Some read, as in the shorter recension, “grace.”}

\xsection{Chapter I.—Praise of the Ephesians.}

\textsc{Shorter.}~\textsc{I have} become acquainted with your name, much-beloved in God, which ye have acquired by the habit of righteousness, according to the faith and love in Jesus Christ our Saviour. Being the followers\footnote{Literally, “imitators;” comp. Eph. v. 1.} of God, and stirring up\footnote{Comp. in the Greek, 2 Tim. i. 6.} yourselves by the blood of God, ye have perfectly accomplished the work which was beseeming to you. For, on hearing that I came bound from Syria for the common name and hope, trusting through your prayers to be permitted to fight with beasts at Rome, that so by martyrdom I may indeed become the disciple of Him “who gave Himself for us, an offering and sacrifice to God,”\footnote{Eph. v. 2.} [ye hastened to see me\footnote{This is wanting in the Greek.}]. I received, therefore,\footnote{Literally, “since therefore,” without any apodosis.} your whole multitude in the name of God, through Onesimus, a man of inexpressible love,\footnote{Or, “unspeakably beloved.”} and your bishop in the flesh, whom I pray you by Jesus Christ to love, and that you would all seek to be like him. And blessed be He who has granted unto you, being worthy, to obtain such an excellent bishop.

\textsc{Longer.}~\textsc{I have} become acquainted with your greatly-desired name in God, which ye have acquired by the habit of righteousness, according to the faith and love in Christ Jesus our Saviour. Being the followers\footnote{Literally, “imitators;” comp. Eph. v. 1.} of the love of God towards man, and stirring up\footnote{Comp. in the Greek, 2 Tim. i. 6.} yourselves by the blood of Christ, you have perfectly accomplished the work which was beseeming to you. For, on hearing that I came bound from Syria for the sake of Christ, our common hope, trusting through your prayers to be permitted to fight with beasts at Rome, that so by martyrdom I may indeed become the disciple of Him “who gave Himself for us, an offering and a sacrifice to God,”\footnote{Eph. v. 2.} [ye hastened to see me\footnote{This is wanting in the Greek.}]. I have therefore received your whole multitude in the name of God, through Onesimus, a man of inexpressible love,\footnote{Or, “unspeakably beloved.”} and who is your bishop, whom I pray you by Jesus Christ to love, and that you would all seek to be like him. Blessed be God, who has granted unto you, who are yourselves so excellent, to obtain such an excellent bishop.

\xsection{Chapter II.—Congratulations and entreaties.}

\textsc{Shorter.}~As to my fellow-servant Burrhus, your deacon in regard to God and blessed in all things,\footnote{Or, “our most blessed deacon in all things pertaining to God.”} I beg that he may continue longer, both for your honour and that of your bishop. And Crocus also, worthy both of God and you, whom I have received as the manifestation\footnote{Literally, “pattern.”} of your love, hath in all things refreshed\footnote{Comp. 1 Cor. xvi. 18, etc.} me, as the Father of our Lord Jesus Christ shall also refresh\footnote{Comp. 1 Cor. xvi. 18, etc.} him; together with Onesimus, and Burrhus, and Euplus, and Fronto, by means of whom, I have, as to love, beheld all of you. May I always have joy of you, if indeed I be worthy of it. It is therefore befitting that you should in every way glorify Jesus Christ, who hath glorified you, that by a unanimous obedience “ye may be perfectly joined together in the same mind, and in the same judgment, and may all speak the same thing concerning the same thing,”\footnote{1 Cor. i. 10.} and that, being subject to the bishop and the presbytery, ye may in all respects be sanctified.

\textsc{Longer.}~As to our fellow-servant Burrhus, your deacon in regard to God and blessed in all things, I pray that he may continue blameless for the honour of the Church, and of your most blessed bishop. Crocus also, worthy both of God and you, whom we have received as the manifestation\footnote{Literally, “pattern.”} of your love to us, hath in all things refreshed\footnote{Comp. 1 Cor. xvi. 18, etc.} me, and “hath not been ashamed of my chain,”\footnote{Comp. 2 Tim. i. 16.} as the Father of our Lord Jesus Christ will also refresh\footnote{Comp. 1 Cor. xvi. 18, etc.} him; together with Onesimus, and Burrhus, and Euplus, and Fronto, by means of whom I have, as to love, beheld all of you. May I always have joy of you, if indeed I be worthy of it. It is therefore befitting that you should in every way glorify Jesus Christ, who hath glorified you, that by a unanimous obedience “ye may be perfectly joined together in the same mind and in the same judgment, and may all speak the same thing concerning the same thing,”\footnote{1 Cor. i. 10.} and that, being subject to the bishop and the presbytery, ye may in all respects be sanctified.

\xsection{Chapter III.—Exhortations to unity.}

\textsc{Shorter.}~I do not issue orders to you, as if I were some great person. For though I am bound for the name [of Christ], I am not yet perfect in Jesus Christ. For now I begin to be a disciple, and I speak to you as fellow-disciples with me. For it was needful for me to have been stirred up by you in faith, exhortation, patience, and long-suffering. But inasmuch as love suffers me not to be silent in regard to you, I have therefore taken\footnote{Comp. Philem. 8, 9.} upon me first to exhort you that ye would all run together in accordance with the will of God. For even Jesus Christ, our inseparable life, is the [manifested] will of the Father; as also bishops, settled everywhere to the utmost bounds [of the earth], are so by the will of Jesus Christ.

\textsc{Longer.}~I do not issue orders to you, as if I were some great person. For though I am bound for His name, I am not yet perfect in Jesus Christ. For now I begin to be a disciple, and I speak to you as my fellow-servants. For it was needful for me to have been admonished by you in faith, exhortation, patience, and long-suffering. But inasmuch as love suffers me not to be silent in regard to you, I have therefore taken\footnote{Comp. Philem. 8, 9.} upon me first to exhort you that ye would run together in accordance with the will of God. For even Jesus Christ does all things according to the will of the Father, as He Himself declares in a certain place, “I do always those things that please Him.”\footnote{John viii. 29.} Wherefore it behoves us also to live according to the will of God in Christ, and to imitate Him as Paul did. For, says he, “Be ye followers of me, even as I also am of Christ.”\footnote{1 Cor. xi. 1.}

\xsection{Chapter IV.—The same continued.}

\textsc{Shorter.}~Wherefore it is fitting that ye should run together in accordance with the will of your bishop, which thing also ye do. For your justly renowned presbytery, worthy of God, is fitted as exactly to the bishop as the strings are to the harp. Therefore in your  concord and harmonious love, Jesus Christ is sung. And do ye, man by man, become a choir, that being harmonious in love, and taking up the song of God in unison, ye may with one voice sing to the Father through Jesus Christ, so that He may both hear you, and perceive by your works that ye are indeed the members of His Son. It is profitable, therefore, that you should live in an unblameable unity, that thus ye may always enjoy communion with God.

\textsc{Longer.}~Wherefore it is fitting that ye also should run together in accordance with the will of the bishop who by God’s appointment\footnote{Literally, “according to God.”} rules over you. Which thing ye indeed of yourselves do, being instructed by the Spirit. For your justly-renowned presbytery, being worthy of God, is fitted as exactly to the bishop as the strings are to the harp. Thus, being joined together in concord and harmonious love, of which Jesus Christ is the Captain and Guardian, do ye, man by man, become but one choir; so that, agreeing together in concord, and obtaining\footnote{Literally, “receiving a union to God in oneness.”} a perfect unity with God, ye may indeed be one in harmonious feeling with God the Father, and His beloved Son Jesus Christ our Lord. For, says He, “Grant unto them, Holy Father, that as I and Thou are one, they also may be one in us.”\footnote{John xvii. 11, 12.} It is therefore profitable that you, being joined together with God in an unblameable unity, should be the followers of the example of Christ, of whom also ye are members.

\xsection{Chapter V.—The praise of unity.}

\textsc{Shorter.}~For if I in this brief space of time, have enjoyed such fellowship with your bishop —I mean not of a mere human, but of a spiritual nature—how much more do I reckon you happy who are so joined to him as the Church is to Jesus Christ, and as Jesus Christ is to the Father, that so all things may agree in unity! Let no man deceive himself: if any one be not within the altar, he is deprived of the bread of God. For if the prayer of one or two possesses\footnote{Matt. xviii. 19.} such power, how much more that of the bishop and the whole Church! He, therefore, that does not assemble with the Church, has even\footnote{Or, “already.”} by this manifested his pride, and condemned himself. For it is written, “God resisteth the proud.”\footnote{Prov. iii. 34; Jas. iv. 6; 1 Pet. v. 5.} Let us be careful, then, not to set ourselves in opposition to the bishop, in order that we may be subject to God.

\textsc{Longer.}~For if I, in this brief space of time, have enjoyed such fellowship with your bishop —I mean not of a mere human, but of a spiritual nature—how much more do I reckon you happy, who so depend\footnote{Some read, “mixed up with.”} on him as the Church does on the Lord Jesus, and the Lord does on God and His Father, that so all things may agree in unity! Let no man deceive himself: if any one be not within the altar, he is deprived of the bread of God. For if the prayer of one or two possesses\footnote{Matt. xviii. 19.} such power that Christ stands in the midst of them, how much more will the prayer of the bishop and of the whole Church, ascending up in harmony to God, prevail for the granting of all their petitions in Christ! He, therefore, that separates himself from such, and does not meet in the society where sacrifices\footnote{Literally, “in the assembly of sacrifices.”} are offered, and with “the Church of the first-born whose names are written in heaven,” is a wolf in sheep’s clothing,\footnote{Matt. vii. 15.} while he presents a mild outward appearance. Do ye, beloved, be careful to be subject to the bishop, and the presbyters and the deacons. For he that is subject to these is obedient to Christ, who has appointed them; but he that is disobedient to these is disobedient to Christ Jesus. And “he that obeyeth not\footnote{Or, “believeth not” (John iii. 36).} the Son shall not see life, but the wrath of God abideth on him.” For he that yields not obedience to his superiors is self-confident, quarrelsome, and proud. But “God,” says [the Scripture] “resisteth the proud, but giveth grace to the humble;”\footnote{Prov. iii. 34; Jas. iv. 6; 1 Pet. v. 5.} and, “The proud have greatly transgressed.” The Lord also says to the priests, “He that heareth you, heareth Me; and he that heareth Me, heareth the Father that sent Me. He that despiseth you, despiseth Me; and he that despiseth Me, despiseth Him that sent Me.”

\xsection{Chapter VI.—Have respect to the bishop as to Christ Himself.}

\textsc{Shorter.}~Now the more any one sees the bishop keeping silence,\footnote{That is, “showing forbearance.”} the more ought he to revere him. For we ought to receive every one whom the Master of the house sends to be over His household,\footnote{Comp. Matt. xxiv. 25.} as we would do Him that sent him. It is manifest, therefore, that we should look upon the bishop even as we would upon the  Lord Himself. And indeed Onesimus himself greatly commends your good order in God, that ye all live according to the truth, and that no sect\footnote{Or, “heresy.”} has any dwelling-place among you. Nor, indeed, do ye hearken to any one rather than to Jesus Christ speaking in truth.

\textsc{Longer.}~The more, therefore, you see the bishop silent, the more do you reverence him. For we ought to receive every one whom the Master of the house sends to be over His household,\footnote{Comp. Matt. xxiv. 25.} as we would do Him that sent him. It is manifest, therefore, that we should look upon the bishop even as we would look upon the Lord Himself, standing, as he does, before the Lord. For “it behoves the man who looks carefully about him, and is active in his business, to stand before kings, and not to stand before slothful men.”\footnote{Prov. xxii. 29, after LXX.} And indeed Onesimus himself greatly commends your good order in God, that ye all live according to the truth, and that no sect has any dwelling-place among you. Nor indeed do ye hearken to any one rather than to Jesus Christ, the true Shepherd and Teacher. And ye are, as Paul wrote to you, “one body and one spirit, because ye have also been called in one hope of the faith.\footnote{Eph. iv. 4.} Since also “there is one Lord, one faith, one baptism, one God and Father of all, who is over all, and through all, and in all.”\footnote{Eph. iv. 5, 6.} Such, then, are ye, having been taught by such instructors, Paul the Christ-bearer, and Timothy the most faithful.

\xsection{Chapter VII.—Beware of false teachers.}

\textsc{Shorter.}~For some are in the habit of carrying about the name [of Jesus Christ] in wicked guile, while yet they practise things unworthy of God, whom ye must flee as ye would wild beasts. For they are ravening dogs, who bite secretly, against whom ye must be on your guard, inasmuch as they are men who can scarcely be cured. There is one Physician who is possessed both of flesh and spirit; both made and not made; God existing in flesh; true life in death; both of Mary and of God; first possible and then impossible,—\footnote{This clause is wanting in the Greek, and has been supplied from the ancient Latin version.} even Jesus Christ our Lord.

\textsc{Longer.}~But some most worthless persons are in the habit of carrying about the name [of Jesus Christ] in wicked guile, while yet they practise things unworthy of God, and hold opinions contrary to the doctrine of Christ, to their own destruction, and that of those who give credit to them, whom you must avoid as ye would wild beasts. For “the righteous man who avoids them is saved for ever; but the destruction of the ungodly is sudden, and a subject of rejoicing.”\footnote{Prov. x. 25, Prov. xi. 3.} For “they are dumb dogs, that cannot bark,”\footnote{Isa. lvi. 10} raving mad, and biting secretly, against whom ye must be on your guard, since they labour under an incurable disease. But our Physician is the only true God, the unbegotten and unapproachable, the Lord of all, the Father and Begetter of the only-begotten Son. We have also as a Physician the Lord our God, Jesus the Christ, the only-begotten Son and Word, before time began,\footnote{Or, “before the ages.”} but who afterwards became also man, of Mary the virgin. For “the Word was made flesh.”\footnote{John i. 14.} Being incorporeal, He was in the body; being impassible, He was in a passible body; being immortal, He was in a mortal body; being life, He became subject to corruption, that He might free our souls from death and corruption, and heal them, and might restore them to health, when they were diseased with ungodliness and wicked lusts.

\xsection{Chapter VIII.—Renewed praise of the Ephesians.}

\textsc{Shorter.}~Let not then any one deceive you, as indeed ye are not deceived, inasmuch as ye are wholly devoted to God. For since there is no strife raging among you which might distress you, ye are certainly living in accordance with God’s will. I am far inferior to you, and require to be sanctified by your Church of Ephesus, so renowned throughout the world. They that are carnal cannot do those things which are spiritual, nor they that are spiritual the things which are carnal; even as faith cannot do the works of unbelief,  nor unbelief the works of faith. But even those things which ye do according to the flesh are spiritual; for ye do all things in Jesus Christ.

\textsc{Longer.}~Let not then any one deceive you, as indeed ye are not deceived; for ye are wholly devoted to God. For when there is no evil desire within you, which might defile and torment you, then do ye live in accordance with the will of God, and are [the servants] of Christ. Cast ye out that which defiles\footnote{It is difficult to translate \textgreek{περίψημα} in this and similar passages; comp. 1 Cor. iv. 13.} you, who are of the\footnote{Literally, “and the.”} most holy Church of the Ephesians, which is so famous and celebrated throughout the world. They that are carnal cannot do those things which are spiritual, nor they that are spiritual the things which are carnal; even as faith cannot do the works of unbelief, nor unbelief the works of faith. But ye, being full of the Holy Spirit, do nothing according to the flesh, but all things according to the Spirit. Ye are complete in Christ Jesus, “who is the Saviour of all men, specially of them that believe.”\footnote{1 Tim. iv. 10.}

\xsection{Chapter IX.—Ye have given no heed to false teachers.}

\textsc{Shorter.}~Nevertheless, I have heard of some who have passed on from this to you, having false doctrine, whom ye did not suffer to sow among you, but stopped your ears, that ye might not receive those things which were sown by them, as being stones\footnote{Comp. 1 Pet. ii. 5.} of the temple of the Father, prepared for the building of God the Father, and drawn up on high by the instrument of Jesus Christ, which is the cross,\footnote{Comp. John xii. 32.} making use of the Holy Spirit as a rope, while your faith was the means by which you ascended, and your love the way which led up to God. Ye, therefore, as well as all your fellow-travellers, are God-bearers, temple-bearers, Christ-bearers, bearers of holiness, adorned in all respects with the commandments of Jesus Christ, in whom also I exult that I have been thought worthy, by means of this Epistle, to converse and rejoice with you, because with respect to your Christian life\footnote{Literally, “according to the other life.”} ye love nothing but God only.

\textsc{Longer.}~Nevertheless, I have heard of some who have passed in among you, holding the wicked doctrine of the strange and evil spirit; to whom ye did not allow entrance to sow their tares, but stopped your ears that ye might not receive that error which was proclaimed by them, as being persuaded that that spirit which deceives the people does not speak the things of Christ, but his own, for he is a lying spirit. But the Holy Spirit does not speak His own things, but those of Christ, and that not from himself, but from the Lord; even as the Lord also announced to us the things that He received from the Father. For, says He, “the word which ye hear is not Mine, but the Father’s, who sent Me.”\footnote{John xiv. 24.} And says He of the Holy Spirit, “He shall not speak of Himself, but whatsoever things He shall hear from Me.”\footnote{John xvi. 13.} And He says of Himself to the Father, “I have,” says He, “glorified Thee upon the earth; I have finished the work which, Thou gavest Me; I have manifested Thy name to men.”\footnote{John xvii. 4, 6.} And of the Holy Ghost, “He shall glorify Me, for He receives of Mine.”\footnote{John xvi. 14.} But the spirit of deceit preaches himself, and speaks his own things, for he seeks to please himself. He glorifies himself, for he is full of arrogance. He is lying, fraudulent, soothing, flattering, treacherous, rhapsodical, trifling, inharmonious, verbose, sordid, and timorous. From his power Jesus Christ will deliver you, who has founded you upon the rock, as being chosen stones, well fitted for the divine edifice of the Father, and who are raised up on high by Christ, who was crucified for you, making use of the Holy Spirit as a rope, and being borne up by faith, while exalted by love from earth to heaven, walking in company with those that are undefiled. For, says [the Scripture], “Blessed are the undefiled in the way, who walk in the law of the Lord.”\footnote{Ps. cxix. 1.} Now the way is unerring, namely, Jesus Christ. For, says He, “I am the way and the life.”\footnote{John xiv. 6.} And this way leads to the Father. For “no man,” says He, “cometh to the Father but by Me.”\footnote{John xiv. 6.} Blessed, then, are ye who are God-bearers, spirit-bearers, temple-bearers, bearers of holiness, adorned in all respects with the commandments of Jesus Christ, being “a royal priesthood, a holy nation, a peculiar people,”\footnote{1 Pet. ii. 9.} on whose account I rejoice exceedingly, and have had the privilege, by this Epistle, of conversing with “the saints which are at Ephesus, the faithful in Christ Jesus.”\footnote{Eph. i. 1.} I rejoice, therefore, over you, that ye do not give heed to vanity, and love nothing according to the flesh, but according to God.

\xsection{Chapter X.—Exhortations to prayer, humility, etc.}

\textsc{Shorter.}~And pray ye without ceasing in behalf of other men. For there is in them hope of repentance that they  may attain to God. See,\footnote{Literally, “permit.”} then, that they be instructed by your works, if in no other way. Be ye meek in response to their wrath, humble in opposition to their boasting: to their blasphemies return\footnote{The verb is here omitted in the original.} your prayers; in contrast to their error, be ye stedfast\footnote{Comp. Col. i. 23.} in the faith; and for their cruelty, manifest your gentleness. While we take care not to imitate their conduct, let us be found their brethren in all true kindness; and let us seek to be followers of the Lord (who ever more unjustly treated, more destitute, more condemned?), that so no plant of the devil may be found in you, but ye may remain in all holiness and sobriety in Jesus Christ, both with respect to the flesh and spirit.

\textsc{Longer.}~And pray ye without ceasing in behalf of other men; for there is hope of the repentance, that they may attain to God. For “cannot he that falls arise again, and he that goes astray return?”\footnote{Jer. viii. 4.} Permit them, then, to be instructed by you. Be ye therefore the ministers of God, and the mouth of Christ. For thus saith the Lord, “If ye take forth the precious from the vile, ye shall be as my mouth.”\footnote{Jer. xv. 19.} Be ye humble in response to their wrath; oppose to their blasphemies your earnest prayers; while they go astray, stand ye stedfast in the faith. Conquer ye their harsh temper by gentleness, their passion by meekness. For “blessed are the meek;”\footnote{Matt. v. 4.} and Moses was meek above all men;\footnote{Num. xii. 3.} and David was exceeding meek.\footnote{Ps. cxxxi. 2.} Wherefore Paul exhorts as follows: “The servant of the Lord must not strive, but be gentle towards all men, apt to teach, patient, in meekness instructing those that oppose themselves.”\footnote{2 Tim. ii. 24, 25.} Do not seek to avenge yourselves on those that injure you, for says [the Scripture], “If I have returned evil to those who returned evil to me.”\footnote{Ps. vii. 4.} Let us make them brethren by our kindness. For say ye to those that hate you, Ye are our brethren, that the name of the Lord may be glorified. And let us imitate the Lord, “who, when He was reviled, reviled not again;”\footnote{1 Pet. ii. 23.} when He was crucified, He answered not; “when He suffered, He threatened not;”\footnote{1 Pet. ii. 23.} but prayed for His enemies, “Father, forgive them; they know not what they do.”\footnote{Luke xxiii. 34.} If any one, the more he is injured, displays the more patience, blessed is he. If any one is defrauded, if any one is despised, for the name of the Lord, he truly is the servant of Christ. Take heed that no plant of the devil be found among you, for such a plant is bitter and salt. “Watch ye, and be ye sober,”\footnote{1 Pet. iv. 7.} in Christ Jesus.

\xsection{Chapter XI.—An exhortation to fear God, etc.}

\textsc{Shorter.}~The last times are come upon us. Let us therefore be of a reverent spirit, and fear the long-suffering of God, that it tend not to our condemnation. For let us either stand in awe of the wrath to come, or show regard for the grace which is at present displayed— one of two things. Only [in one way or another] let us be found in Christ Jesus unto the true life. Apart from Him, let nothing attract\footnote{Literally, “let nothing become you.”} you, for whom I bear about these bonds, these spiritual jewels, by which may I arise through your prayers, of which I entreat I may always be a partaker, that I may be found in the lot of the Christians of Ephesus, who have always been of the same mind with the apostles through the power of Jesus Christ.

\textsc{Longer.}~The last times are come upon us. Let us therefore be of a reverent spirit, and fear the long-suffering of God, lest we despise the riches of His goodness and forbearance.\footnote{Rom. ii. 4.} For let us either fear the wrath to come, or let us love the present joy in the life that now is; and let our present and true joy be only this, to be found in Christ Jesus, that we may truly live. Do not at any time desire so much as even to breathe apart from Him. For He is my hope; He is my boast; He is my never-failing riches, on whose account I bear about with me these bonds from Syria to Rome, these spiritual jewels, in which may I be perfected through your prayers, and become a partaker of the sufferings of Christ, and have fellowship with Him in His death, His resurrection from the dead, and His everlasting life.\footnote{Phil. iii. 10.} May I attain to this, so that I may be found in the lot of the Christians of Ephesus, who have always had intercourse with the apostles by the power of Jesus Christ, with Paul, and John, and Timothy the most faithful.

\xsection{Chapter XII.—Praise of the Ephesians.}

\textsc{Shorter.}~I know both who I am, and to whom I write. I am a condemned man, ye have been the objects of  mercy; I am subject to danger, ye are established in safety. Ye are the persons through\footnote{Literally, “ye are the passage of.”} whom those pass that are cut off for the sake of God. Ye are initiated into the mysteries of the Gospel with Paul, the holy, the martyred, the deservedly most happy, at whose feet\footnote{Literally, “footsteps.”} may I be found, when I shall attain to God; who in all his Epistles makes mention of you in Christ Jesus.

\textsc{Longer.}~I know both who I am, and to whom I write. I am the very insignificant Ignatius, who have my lot with\footnote{Literally, “am like to.”} those who are exposed to danger and condemnation. But ye have been the objects of mercy, and are established in Christ. I am one delivered over [to death], but the least of all those that have been cut off for the sake of Christ, “from the blood of righteous Abel”\footnote{Matt. xxiii. 35.} to the blood of Ignatius. Ye are initiated into the mysteries of the Gospel with Paul, the holy, the martyred, inasmuch as he was “a chosen vessel;”\footnote{Acts ix. 15.} at whose feet may I be found, and at the feet of the rest of the saints, when I shall attain to Jesus Christ, who is always mindful of you in His prayers.

\xsection{Chapter XIII.—Exhortation to meet together frequently for the worship of God.}

\textsc{Shorter.}~Take heed, then, often to come together to give thanks to God, and show forth His praise. For when ye assemble frequently in the same place, the powers of Satan are destroyed, and the destruction at which he aims\footnote{Literally, “his destruction.”} is prevented by the unity of your faith. Nothing is more precious than peace, by which all war, both in heaven and earth,\footnote{Literally, “of heavenly and earthly things.”} is brought to an end.

\textsc{Longer.}~Take heed, then, often to come together to give thanks to God, and show forth His praise. For when ye come frequently together in the same place, the powers of Satan are destroyed, and his “fiery darts”\footnote{Eph. vi. 16.} urging to sin fall back ineffectual. For your concord and harmonious faith prove his destruction, and the torment of his assistants. Nothing is better than that peace which is according to Christ, by which all war, both of aërial and terrestrial spirits, is brought to an end. “For we wrestle not against blood and flesh, but against principalities and powers, and against the rulers of the darkness of this world, against spiritual wickedness in heavenly places.”\footnote{Eph. vi. 12.}

\xsection{Chapter XIV.—Exhortations to faith and love.}

\textsc{Shorter.}~None of these things is hid from you, if ye perfectly possess that faith and love towards Christ Jesus\footnote{1 Tim. i. 14.} which are the beginning and the end of life. For the beginning is faith, and the end is love.\footnote{1 Tim. i. 5.} Now these two, being inseparably connected together,\footnote{Literally, “being in unity.”} are of God, while all other things which are requisite for a holy life follow after them. No man [truly] making a profession of faith sinneth;\footnote{Comp. 1 John iii. 7.} nor does he that possesses love hate any one. The tree is made manifest by its fruit;\footnote{Matt. xii. 33.} so those that profess themselves to be Christians shall be recognised by their conduct. For there is not now a demand for mere profession,\footnote{Literally, “there is not now the work of profession.”} but that a man be found continuing in the power of faith to the end.

\textsc{Longer.}~Wherefore none of the devices of the devil shall be hidden from you, if, like Paul, ye perfectly possess that faith and love towards Christ\footnote{1 Tim. i. 14.} which are the beginning and the end of life. The beginning of life is faith, and the end is love. And these two being inseparably connected together, do perfect the man of God; while all other things which are requisite to a holy life follow after them. No man making a profession of faith ought to sin, nor one possessed of love to hate his brother. For He that said, “Thou shalt love the Lord thy God,”\footnote{Luke x. 27.} said also, “and thy neighbour as thyself.”\footnote{Luke x. 27.} Those that profess themselves to be Christ’s are known not only by what they say, but by what they practise. “For the tree is known by its fruit.”\footnote{Matt. xii. 33.}

\xsection{Chapter XV.—Exhortation to confess Christ by silence as well as speech.}

\textsc{Shorter.}~It is better for a man to be silent and be [a Christian], than to talk and not to be one. It is good to teach, if he who speaks also acts. There is then one Teacher, who spake and it was done; while even those things which He did in silence are worthy of  the Father. He who possesses the word of Jesus, is truly able to hear even His very silence, that he may be perfect, and may both act as he speaks, and be recognised by his silence. There is nothing which is hid from God, but our very secrets are near to Him. Let us therefore do all things as those who have Him dwelling in us, that we may be His temples,\footnote{1 Cor. vi. 19.} and He may be in us as our God, which indeed He is, and will manifest Himself before our faces. Wherefore we justly love Him.

\textsc{Longer.}~It is better for a man to be silent and be [a Christian], than to talk and not to be one. “The kingdom of God is not in word, but in power.”\footnote{1 Cor. iv. 20.} Men “believe with the heart, and confess with the mouth,” the one “unto righteousness,” the other “unto salvation.”\footnote{Rom. x. 10.} It is good to teach, if he who speaks also acts. For he who shall both “do and teach, the same shall be great in the kingdom.”\footnote{Matt. v. 19.} Our Lord and God, Jesus Christ, the Son of the living God, first did and then taught, as Luke testifies, “whose praise is in the Gospel through all the Churches.”\footnote{2 Cor. viii. 18.} There is nothing which is hid from the Lord, but our very secrets are near to Him. Let us therefore do all things as those who have Him dwelling in us, that we may be His temples,\footnote{1 Cor. vi. 19.} and He may be in us as God. Let Christ speak in us, even as He did in Paul. Let the Holy Spirit teach us to speak the things of Christ in like manner as He did.

\xsection{Chapter XVI.—The fate of false teachers.}

\textsc{Shorter.}~Do not err, my brethren.\footnote{Comp. Jas. i. 16.} Those that corrupt families shall not inherit the kingdom of God.\footnote{1 Cor. vi. 9, 10.} If, then, those who do this as respects the flesh have suffered death, how much more shall this be the case with any one who corrupts by wicked doctrine the faith of God, for which Jesus Christ was crucified! Such an one becoming defiled [in this way], shall go away into everlasting fire, and so shall every one that hearkens unto him.

\textsc{Longer.}~Do not err, my brethren.\footnote{Comp. Jas. i. 16.} Those that corrupt families shall not inherit the kingdom of God.\footnote{1 Cor. vi. 9, 10.} And if those that corrupt mere human families are condemned to death, how much more shall those suffer everlasting punishment who endeavour to corrupt the Church of Christ, for which the Lord Jesus, the only-begotten Son of God, endured the cross, and submitted to death! Whosoever, “being waxen fat,”\footnote{Deut. xxxii. 15.} and “become gross,” sets at nought His doctrine, shall go into hell. In like manner, every one that has received from God the power of distinguishing, and yet follows an unskilful shepherd, and receives a false opinion for the truth, shall be punished. “What communion hath light with darkness? or Christ with Belial? Or what portion hath he that believeth with an infidel? or the temple of God with idols?”\footnote{2 Cor. vi. 14–16.} And in like manner say I, what communion hath truth with falsehood? or righteousness with unrighteousness? or true doctrine with that which is false?

\xsection{Chapter XVII.—Beware of false doctrines.}

\textsc{Shorter.}~For this end did the Lord suffer the ointment to be poured upon His head,\footnote{Comp. John xii. 7.} that He might breathe immortality into His Church. Be not ye anointed with the bad odour of the doctrine of the prince of this world; let him not lead you away captive from the life which is set before you. And why are we not all prudent, since we have received the knowledge of God, which is Jesus Christ? Why do we foolishly perish, not recognising the gift which the Lord has of a truth sent to us?

\textsc{Longer.}~For this end did the Lord suffer the ointment to be poured upon His head,\footnote{Comp. John xii. 7.} that His Church might breathe forth immortality. For saith [the Scripture], “Thy name is as ointment poured forth; therefore have the virgins loved Thee; they have drawn Thee; at the odour of Thine ointments we will run after Thee.”\footnote{Cant. i. 3, 4.} Let no one be anointed with the bad odour of the doctrine of [the prince of] this world; let not the holy Church of God be led captive by his subtlety, as was the first woman.\footnote{Literally, “before the ages.”} Why do we not, as gifted with reason, act wisely? When we had received from Christ, and had grafted in us the faculty of judging concerning God, why do we fall headlong into ignorance? and why, through a careless neglect of acknowledging the gift which we have received, do we foolishly perish?

\xsection{Chapter XVIII.—The glory of the cross.}

\textsc{Shorter.}~Let my spirit be counted as nothing\footnote{Again, \textgreek{περίψημα}, translated “offscouring,” 1 Cor. iv. 13.} for the sake of the cross, which is a stumbling-block\footnote{Comp. 1 Cor. i. 18.} to those that do not believe, but to us salvation and  life eternal. “Where is the wise man? where the disputer?”\footnote{1 Cor. i. 20.} Where is the boasting of those who are styled prudent? For our God, Jesus Christ, was, according to the appointment\footnote{Or, “economy,” or “dispensation.” Comp. Col. i. 25; 1 Tim. i. 4.} of God, conceived in the womb by Mary, of the seed of David, but by the Holy Ghost. He was born and baptized, that by His passion He might purify the water.

\textsc{Longer.}~The cross of Christ is indeed a stumbling-block to those that do not believe, but to the believing it is salvation and life eternal. “Where is the wise man? where the disputer?”\footnote{1 Cor. i. 20.} Where is the boasting of those who are called mighty? For the Son of God, who was begotten before time began,\footnote{Literally, “before the ages.”} and established all things according to the will of the Father, He was conceived in the womb of Mary, according to the appointment of God, of the seed of David, and by the Holy Ghost. For says [the Scripture], “Behold, a virgin shall be with child, and shall bring forth a son, and He shall be called Immanuel.”\footnote{Isa. vii. 14; Matt. i. 23.} He was born and was baptized by John, that He might ratify the institution committed to that prophet.

\xsection{Chapter XIX.—Three celebrated mysteries.}

\textsc{Shorter.}~Now the virginity of Mary was hidden from the prince of this world, as was also her offspring, and the death of the Lord; three mysteries of renown,\footnote{Literally, “of noise.”} which were wrought in silence by\footnote{Or, “in the silence of God”—divine silence.} God. How, then, was He manifested to the world?\footnote{Literally, “to the ages.”} A star shone forth in heaven above all the other stars, the light of which was inexpressible, while its novelty struck men with astonishment. And all the rest of the stars, with the sun and moon, formed a chorus to this star, and its light was exceedingly great above them all. And there was agitation felt as to whence this new spectacle came, so unlike to everything else [in the heavens]. Hence every kind of magic was destroyed, and every bond of wickedness disappeared; ignorance was removed, and the old kingdom abolished, God Himself being manifested in human form for the renewal of eternal life. And now that took a beginning which had been prepared by God. Henceforth all things were in a state of tumult, because He meditated the abolition of death.

\textsc{Longer.}~Now the virginity of Mary was hidden from the prince of this world, as was also her offspring, and the death of the Lord; three mysteries of renown,\footnote{Literally, “of noise.”} which were wrought in silence, but have been revealed to us. A star shone forth in heaven above all that were before it, and its light was inexpressible, while its novelty struck men with astonishment. And all the rest of the stars, with the sun and moon, formed a chorus to this star. It far exceeded them all in brightness, and agitation was felt as to whence this new spectacle [proceeded]. Hence worldly wisdom became folly; conjuration was seen to be mere trifling; and magic became utterly ridiculous. Every law\footnote{Some read, “bond.”} of wickedness vanished away; the darkness of ignorance was dispersed; and tyrannical authority was destroyed, God being manifested as a man, and man displaying power as God. But neither was the former a mere imagination,\footnote{Literally, “opinion.”} nor did the second imply a bare humanity;\footnote{Literally, “bareness.”} but the one was absolutely true,\footnote{Literally, “truth.”} and the other an economical arrangement.\footnote{Literally, “an economy.”} Now that received a beginning which was perfected by God.\footnote{Or, “that which was perfect received a beginning from God.”} Henceforth all things were in a state of tumult, because He meditated the abolition of death.

\xsection{Chapter XX.—Promise of another letter.}

If Jesus Christ shall graciously permit me through your prayers, and if it be His will, I shall, in a second little work which I will write to you, make further manifest to you [the nature of] the dispensation of which I have begun [to treat], with respect to the new man, Jesus Christ, in His faith and in His love, in His suffering and in His resurrection. Especially [will I do this\footnote{The punctuation and meaning are here doubtful.}] if the Lord make known to me that ye come together  man by man in common through grace, individually,\footnote{Literally, “by name.”} in one faith, and in Jesus Christ, who was of the seed of David according to the flesh, being both the Son of man and the Son of God, so that ye obey the bishop and the presbytery with an undivided mind, breaking one and the same bread, which is the medicine of immortality, and the antidote to prevent us from dying, but [which causes] that we should live for ever in Jesus Christ.

\xsection{Chapter XX.—Exhortations to stedfastness and unity.}

Stand fast, brethren, in the faith of Jesus Christ, and in His love, in His passion, and in His resurrection. Do ye all come together in common, and individually,\footnote{Literally, “by name.”} through grace, in one faith of God the Father, and of Jesus Christ His only-begotten Son, and “the first-born of every creature,”\footnote{Col. i. 15.} but of the seed of David according to the flesh, being under the guidance of the Comforter, in obedience to the bishop and the presbytery with an undivided mind, breaking one and the same bread, which is the medicine of immortality, and the antidote which prevents us from dying, but a cleansing remedy driving away evil, [which causes] that we should live in God through Jesus Christ.

\xsection{Chapter XXI.—Conclusion.}

\textsc{Shorter.}~My soul be for yours and theirs\footnote{Some render, “May I, in my turn, be the means of refreshing you and those,” etc.} whom, for the honour of God, ye have sent to Smyrna; whence also I write to you, giving thanks unto the Lord, and loving Polycarp even as I do you. Remember me, as Jesus Christ also remembered you. Pray ye for the Church which is in Syria, whence I am led bound to Rome, being the last of the faithful who are there, even as I have been thought worthy to be chosen\footnote{Literally, “to be found for.”} to show forth the honour of God. Farewell in God the Father, and in Jesus Christ, our common hope.

\textsc{Longer.}~My soul be for yours and theirs\footnote{Some render, “May I, in my turn, be the means of refreshing you and those,” etc.} whom, for the honour of God, ye have sent to Smyrna; whence also I write to you, giving thanks to the Lord, and loving Polycarp even as I do you. Remember me, as Jesus Christ also remembers you, who is blessed for evermore. Pray ye for the Church of Antioch which is in Syria, whence I am led bound to Rome, being the last of the faithful that are there, who\footnote{Some read, “even as.”} yet have been thought worthy to carry these chains to the honour of God. Fare ye well in God the Father, and the Lord Jesus Christ, our common hope, and in the Holy Ghost. Fare ye well. Amen. Grace [be with you].\footnote{Some omit, “Grace [be with you].”}

\xchapter{Epistle to the Magnesians}

\textsc{Shorter.}~\emph{Ignatius, who is also called Theophorus, to the [Church] blessed in the grace of God the Father, in Jesus Christ our Saviour, in whom I salute the Church which is at Magnesia, near the Mæander, and wish it abundance of happiness in God the Father, and in Jesus Christ.}

\textsc{Longer.}~\emph{Ignatius, who is also called Theophorus, to the [Church] blessed in the grace of God the Father, in Jesus Christ our Saviour, in whom I salute the Church which is at Magnesia, near the Mæander, and wish it abundance of happiness in God the Father, and in Jesus Christ, our Lord, in whom may you have abundance of happiness.}

\xsection{Chapter I.—Reason of writing the epistle.}

\textsc{Shorter.}~\textsc{Having} been informed of your godly\footnote{Literally, “according to God.”} love, so well-ordered, I rejoiced greatly, and determined to commune with you in the faith of Jesus Christ. For as one who has been thought worthy of the most honourable of all names,\footnote{Literally, “of the most God-becoming name,” referring either to the appellation “Theophorus,” or to that of “martyr” or “confessor.”} in those bonds which I bear about, I commend the Churches, in which I pray for a union both of the flesh and spirit of Jesus Christ, the constant source of our life, and of faith and love, to which nothing is to be preferred, but especially of Jesus and the Father, in whom, if we endure all the assaults of the prince of this world, and escape them, we shall enjoy God.

\textsc{Longer.}~\textsc{Having} been informed of your godly\footnote{Literally, “according to God.”} love, so well-ordered, I rejoiced greatly, and determined to commune with you in the faith of Jesus Christ. For as one who has been thought worthy of a divine and desirable name, in those bonds which I bear about, I commend the Churches, in which I pray for a union both of the flesh and spirit of Jesus Christ, “who is the Saviour of all men, but specially of them that believe;”\footnote{1 Tim. iv. 10.} by whose blood ye were redeemed; by whom ye have known God, or rather have been known by Him;\footnote{Comp. Gal. iv. 9.} in whom enduring, ye shall escape all the assaults of this world: for “He is faithful, who will not suffer you to be tempted above that which ye are able.”\footnote{1 Cor. x. 13.}

\xsection{Chapter II.—I rejoice in your messengers.}

\textsc{Shorter.}~Since, then, I have had the privilege of seeing you, through Damas your most worthy bishop, and through your worthy presbyters Bassus and Apollonius, and through my fellow-servant the deacon Sotio, whose friendship may I ever enjoy, inasmuch as he is subject to the bishop as to the grace of God, and to the presbytery as to the law of Jesus Christ, [I now write\footnote{The apodosis is here wanting in the original, but must evidently be supplied in some such way as above.} to you].

\textsc{Longer.}~Since, then, I have had the privilege of seeing you, through Damas your most worthy\footnote{Literally, “worthy of God.”} bishop, and through your worthy\footnote{Literally, “worthy of God.”} presbyters Bassus and Apollonius, and through my fellow-servant the deacon Sotio, whose friendship may I ever enjoy,\footnote{Literally, “whom may I enjoy.”} inasmuch as he, by the grace of God, is subject to the bishop and presbytery, in the law of Jesus Christ, [I now write\footnote{The apodosis is here wanting in the original, but must evidently be supplied in some such way as above.} to you].

\xsection{Chapter III.—Honour your youthful bishop.}

\textsc{Shorter.}~Now it becomes you also not to treat your bishop too familiarly on account of his youth,\footnote{Literally, “to use the age of your bishop.”} but to yield him all reverence, having respect to\footnote{Literally, “according to.”} the power of God the Father, as I have known even holy presbyters do, not judging rashly, from the manifest youthful appearance\footnote{Literally, “youthful condition.”} [of their bishop], but as being themselves prudent in God, submitting to him, or rather not to him, but to the Father of Jesus Christ, the bishop of us all. It is therefore fitting that you should, after no hypocritical fashion, obey [your bishop], in honour of Him who has willed us [so to do], since he that does not so deceives not [by such conduct] the bishop that is visible, but seeks to mock Him that is invisible. And all such conduct has reference not to man,\footnote{Literally, “to flesh.”} but to God, who knows all secrets.

\textsc{Longer.}~Now it becomes you also not to despise the age of your bishop, but to yield him all reverence, according to the will of God the Father, as I have known even holy presbyters do, not having regard to the manifest youth [of their bishop], but to his knowledge in God; inasmuch as “not the ancient are [necessarily] wise, nor do the aged understand prudence; but there is a spirit in men.”\footnote{Job xxxii. 8, 9.} For Daniel the wise, at twelve years of age, became possessed of the divine Spirit, and convicted the elders, who in vain carried their grey hairs, of being false accusers, and of lusting after the beauty of another man’s wife.\footnote{Susanna (Apoc.).} Samuel also, when he was but a little child, reproved Eli, who was ninety years old, for giving honour to his sons rather than to God.\footnote{1 Sam. iii. 1.} In like manner, Jeremiah also received this message from God, “Say not, I am a child.”\footnote{Jer. i. 7.} Solomon too, and Josiah, [exemplified the same thing.] The former, being made king at twelve years of age, gave that terrible and difficult judgment in the case of the two women concerning their children.\footnote{1 Kings iii. 16.} The latter, coming to the throne when eight years old\footnote{2 Kings xxii., xxiii..} cast down the altars and temples [of the idols], and burned down the groves, for they were dedicated to demons, and not to God. And he slew the false priests, as the corrupters and deceivers of men, and not the worshippers of the Deity. Wherefore youth is not to be despised when it is devoted to God. But he is to be despised who is of a wicked mind, although he be old, and full of wicked days.\footnote{Susanna 52 (Apoc.).} Timothy the Christ-bearer was young, but hear what his teacher writes to him: “Let no man despise thy youth, but be thou an example of the believers in word and in conduct.”\footnote{1 Tim. iv. 12.} It is becoming, therefore, that ye also should be obedient to your bishop, and contradict him in nothing; for it is a fearful thing to contradict any such person. For no one does [by such conduct] deceive him that is visible, but does [in reality] seek to mock Him that is invisible, who, however, cannot be mocked by any one. And every such act has respect not to man, but to God. For God says to Samuel, “They have not mocked thee, but Me.”\footnote{1 Sam. viii. 7.} And Moses declares, “For their murmuring is not against us, but against the Lord God.”\footnote{Ex. xvi. 8.} No one of those has, [in fact,] remained unpunished, who rose up against their superiors. For Dathan and Abiram did not speak against the law, but against Moses,\footnote{Num. xvi. 1.} and were cast down alive into Hades. Korah also,\footnote{Num. xvi. 31.} and the two hundred and fifty who conspired with him against Aaron, were destroyed by fire. Absalom, again,\footnote{2 Sam. xviii. 14.} who had slain his brother, became suspended on a tree, and had his evil-designing heart thrust through with darts. In like manner was Abeddadan\footnote{Sheba is referred to under this name: see 2 Sam. xx. 22.} beheaded for the same reason. Uzziah,\footnote{2 Chron. xxvi. 20.} when he presumed to oppose the priests and the priesthood, was smitten with leprosy. Saul also was dishonoured,\footnote{1 Sam. xiii. 11.} because he did not wait for Samuel the high priest. It behoves you, therefore, also to reverence your superiors.

\xsection{Chapter IV.—Some wickedly act independently of the bishop.}

\textsc{Shorter.}~It is fitting, then, not only to be called Christians, but to be so in reality: as some indeed give one the title of bishop, but do all things without him. Now such persons seem to me to be not possessed of a good conscience, seeing they are not stedfastly gathered together according to the commandment.

\textsc{Longer.}~It is fitting, then, not only to be called Christians, but to be so in reality. For it is not the being called so, but the being really so, that renders a man blessed. To those who indeed talk of the bishop, but do all things without him, will He who is the true and first Bishop, and the only High Priest by nature, declare, “Why call ye Me Lord, and do not the things which I say?”\footnote{Luke vi. 46.} For such persons seem to me not possessed of a good conscience, but to be simply dissemblers and hypocrites.

\xsection{Chapter V.—Death is the fate of all such.}

\textsc{Shorter.}~Seeing, then, all things have an end, these two things are simultaneously set before us—death and life; and every one shall go unto his own place. For as there are two kinds of coins, the one of God, the other of the world, and each of these has its special character stamped upon it, [so is it also here.]\footnote{The apodosis is wanting in the original, and some prefer finding it in the following sentence.} The unbelieving are of this world; but the believing have, in love, the character of God the Father by Jesus Christ, by whom, if we are not in readiness to die into His passion,\footnote{Or, “after the likeness of His passion.”} His life is not in us.

\textsc{Longer.}~Seeing, then, all things have an end, and there is set before us life upon our observance [of God’s precepts], but death as the result of disobedience, and every one, according to the choice he makes, shall go to his own place, let us flee from death, and make choice of life. For I remark, that two different characters are found among men—the one true coin, the other spurious. The truly devout man is the right kind of coin, stamped by God Himself. The ungodly man, again, is false coin, unlawful, spurious, counterfeit, wrought not by God, but by the devil. I do not mean to say that there are two different human natures, but that there is one humanity, sometimes belonging to God, and sometimes to the devil. If any one is truly religious, he is a man of God; but if he is irreligious, he is a man of the devil, made such, not by nature, but by his own choice. The unbelieving bear the image of the prince of wickedness. The believing possess the image of their Prince, God the Father, and Jesus Christ, through whom, if we are not in readiness to die for the truth into His passion,\footnote{Or, “after the likeness of His passion.”} His life is not in us.

\xsection{Chapter VI.—Preserve harmony.}

\textsc{Shorter.}~Since therefore I have, in the persons before mentioned, beheld the whole multitude of you in faith and love, I exhort you to study to do all things with a divine harmony,\footnote{Literally, “in harmony of God.”} while your bishop presides in the place of God, and your presbyters in the place of the assembly of the apostles, along with your deacons, who are most dear to me, and are entrusted with the ministry of Jesus Christ, who was with the Father before the beginning of time,\footnote{Literally, “before the ages.”} and in the end was revealed. Do ye all then, imitating the same divine conduct,\footnote{Literally, “receiving the like manners of God.”} pay respect to one another, and let no one look upon his neighbour after the flesh, but do ye continually love each other in Jesus Christ. Let nothing exist among you that may divide you; but be ye united with your bishop, and those that preside over you, as a type and evidence of your immortality.\footnote{The meaning is here doubtful.}

\textsc{Longer.}~Since therefore I have, in the persons before mentioned, beheld the whole multitude of you in faith and love, I exhort you to study to do all things with a divine harmony,\footnote{Literally, “in harmony of God.”} while your bishop presides in the place of God, and your presbyters in the place of the assembly of the apostles, along with your deacons, who are most dear to me, and are entrusted with the ministry of Jesus Christ. He, being begotten by the Father before the beginning of time,\footnote{Literally, “before the ages.”} was God the Word, the only-begotten Son, and remains the same for ever; for “of His kingdom there shall be no end,”\footnote{Dan. ii. 44, Dan. vii. 14, 27.} says Daniel the prophet. Let us all therefore love one another in harmony, and let no one look upon his neighbour according to the flesh, but in Christ Jesus. Let nothing exist among you which may divide you; but be ye united with your bishop, being through him subject to God in Christ.

\xsection{Chapter VII.—Do nothing without the bishop and presbyters.}

\textsc{Shorter.}~As therefore the Lord did nothing without the Father, being united to Him, neither by Himself nor by the apostles, so neither do ye anything without the bishop and presbyters. Neither endeavour that anything appear reasonable and proper to yourselves apart; but being come together into the same place, let there be one prayer, one supplication, one mind, one hope, in love and in joy undefiled. There is one Jesus Christ, than whom nothing is more excellent. Do ye therefore all run together as into one temple of God, as to one altar, as to one Jesus Christ, who came forth from one Father, and is with and has gone to one.

\textsc{Longer.}~As therefore the Lord does nothing without the Father, for says He, “I can of mine own self do nothing,”\footnote{John v. 30.} so do ye, neither presbyter, nor deacon, nor layman, do anything without the bishop. Nor let anything appear commendable to you which is destitute of his approval.\footnote{Or, “contrary to his judgment.”} For every such thing is sinful, and opposed [to the will of] God. Do ye all come together into the same place for prayer. Let there be one common supplication, one mind, one hope, with faith unblameable in Christ Jesus, than which nothing is more excellent. Do ye all, as one man, run together into the temple of God, as unto one altar, to one Jesus Christ, the High Priest of the unbegotten God.

\xsection{Chapter VIII.—Caution against false doctrines.}

\textsc{Shorter.}~Be not deceived with strange doctrines, nor with old fables, which are unprofitable. For if we still live according to the Jewish law, we acknowledge that we have not received grace. For the divinest prophets lived according to Christ Jesus. On this account also they were persecuted, being inspired by His grace to fully convince the unbelieving that there is one God, who has manifested Himself by Jesus Christ His Son, who is His eternal Word, not proceeding forth from silence,\footnote{Some have argued that the Gnostic \textgreek{Σιγή}, \emph{silence}, is here referred to, and have consequently inferred that this epistle could not have been written by Ignatius.} and who in all things pleased Him that sent Him.

\textsc{Longer.}~Be not deceived with strange doctrines, “nor give heed to fables and endless genealogies,”\footnote{1 Tim. i. 4.} and things in which the Jews make their boast. “Old things are passed away: behold, all things have become new.”\footnote{2 Cor. v. 17.} For if we still live according to the Jewish law, and the circumcision of the flesh, we deny that we have received grace. For the divinest prophets lived according to Jesus Christ. On this account also they were persecuted, being inspired by grace to fully convince the unbelieving that there is one God, the Almighty, who has manifested Himself by Jesus Christ His Son, who is His Word, not spoken, but essential. For He is not the voice of an articulate utterance, but a substance begotten by divine power, who has in all things pleased Him that sent Him.\footnote{Some read \textgreek{ὑποστήσαντι}, “that gave Him His \emph{hypostasis}, or substance.”}

\xsection{Chapter IX.—Let us live with Christ.}

\textsc{Shorter.}~If, therefore, those who were brought up in the ancient order of things\footnote{Literally, “in old things.”} have come to the possession of a new\footnote{Or, “newness of.”} hope, no longer observing the Sabbath, but living in the observance\footnote{Or, “according to.”} of the Lord’s Day, on which also our life has sprung up again by Him and by His death—whom some deny, by which mystery we have obtained faith,\footnote{Literally, “we have received to believe.”} and therefore endure, that we may be found the disciples of Jesus Christ, our only Master—how shall we be able to live apart from Him, whose disciples the prophets themselves in the Spirit did wait for Him as their Teacher? And therefore He whom they rightly waited for, being come, raised them from the dead.\footnote{Comp. Matt. xxvii. 52.}

\textsc{Longer.}~If, then, those who were conversant with the ancient Scriptures came to newness of hope, expecting the coming of Christ, as the Lord teaches us when He says, “If ye had believed Moses, ye would have believed Me, for he wrote of Me;”\footnote{John v. 46.} and again, “Your father Abraham rejoiced to see My day, and he saw it, and was glad; for before Abraham was, I am;”\footnote{John viii. 56, 58.} how shall we be able to live without Him? The prophets were His servants, and foresaw Him by the Spirit, and waited for Him as their Teacher, and expected Him as their Lord and Saviour, saying, “He will come and save us.”\footnote{Isa. xxxv. 4.} Let us therefore no longer keep the Sabbath after the Jewish manner, and rejoice in days of idleness; for “he that does not work, let him not eat.”\footnote{2 Thess. iii. 10.} For say the [holy] oracles, “In the sweat of thy face shalt thou eat thy bread.” \footnote{Gen. iii. 19.} But let every one of you keep the Sabbath after a spiritual manner, rejoicing in meditation on the law, not in relaxation of the body, admiring the workmanship of God, and not eating things prepared the day before, nor using lukewarm drinks, and walking within a prescribed space, nor finding delight in dancing and plaudits which have no sense in them.\footnote{Reference is here made to well-known Jewish opinions and practices with respect to the Sabbath. The Talmud fixes 2000 cubits as the space lawful to be traversed. Philo \emph{(De Therap.)} refers to the dancing, etc.} And after the observance of the Sabbath, let every friend of Christ keep the Lord’s Day as a festival, the resurrection-day, the queen and chief of all the days [of the week]. Looking forward to this, the prophet declared, “To the end, for the eighth day,”\footnote{Ps. vi., Ps. xii. (inscrip.). [N.B.—The reference is to the title of these two psalms, as rendered by the LXX. \textgreek{Εἰς τὸ τέλος ὑπὲρ τῆς ὀγδόης}.]} on which our life both sprang up again, and the victory over death was obtained in Christ, whom the children of perdition, the enemies of the Saviour, deny, “whose god is their belly, who mind earthly things,”\footnote{Phil. iii. 18, 19.} who are “lovers of pleasure, and not lovers of God, having a form of godliness, but denying the power thereof.”\footnote{2 Tim. iii. 4.} These make merchandise of Christ, corrupting His word, and giving up Jesus to sale: they are corrupters of women, and covetous of other men’s possessions, swallowing up wealth\footnote{Literally, “whirlpools of wealth.”} insatiably; from whom may ye be delivered by the mercy of God through our Lord Jesus Christ!

\xsection{Chapter X.—Beware of Judaizing.}

\textsc{Shorter.}~Let us not, therefore, be insensible to His kindness. For were He to reward us according to our works, we should cease to be. Therefore, having become His disciples, let us learn to live according to the principles of Christianity.\footnote{Literally, “according to Christianity.”} For whosoever is called by any other name besides this, is not of God. Lay aside, therefore, the evil, the old, the sour leaven, and be ye changed into the new leaven, which is Jesus Christ. Be ye salted in Him, lest any one among you should be corrupted, since by your savour ye shall be convicted. It is absurd to profess\footnote{Some read, “to name.”} Christ Jesus, and to Judaize. For Christianity did not embrace\footnote{Literally, “believe into,” merge into.} Judaism, but Judaism Christianity, that so every tongue which believeth might be gathered together to God.

\textsc{Longer.}~Let us not, therefore, be insensible to His kindness. For were He to reward us according to our works, we should cease to be. For “if Thou, Lord, shalt mark iniquities, O Lord, who shall stand?”\footnote{Ps. cxxx. 3.} Let us therefore prove ourselves worthy of that name which we have received. For whosoever is called by any other name besides this, he is not of God; for he has not received the prophecy which speaks thus concerning us: “The people shall be called by a new name, which the Lord shall name them, and shall be a holy people.”\footnote{Isa. lxii. 2, 12.} This was first fulfilled in Syria; for “the disciples were called Christians at Antioch,”\footnote{Acts xi. 26.} when Paul and Peter were laying the foundations of the Church. Lay aside, therefore, the evil, the old, the corrupt leaven,\footnote{1 Cor. v. 7.} and be ye changed into the new leaven of grace. Abide in Christ, that the stranger\footnote{Or, “enemy.”} may not have dominion over you. It is absurd to speak of Jesus Christ with the tongue, and to cherish in the mind a Judaism which has now come to an end. For where there is Christianity there cannot be Judaism. For Christ is one, in whom every nation that believes, and every tongue that confesses, is gathered unto God. And those that were of a stony heart have become the children of Abraham, the friend of God;\footnote{Matt. iii. 9; Isa. xli. 8; Jas. ii. 23. Some read, “children of God, friends of Abraham.”} and in his seed all those have been blessed\footnote{Gen. xxviii. 14.} who were ordained to eternal life\footnote{Acts xiii. 48.} in Christ.

\xsection{Chapter XI.—I write these things to warn you.}

\textsc{Shorter.}~These things [I address to you], my beloved, not that I know any of you to be in such a state;\footnote{i.e., addicted to the error of Judaizing.} but, as less than any of you, I desire to guard you beforehand, that ye fall not upon the hooks of vain doctrine, but that  ye attain to full assurance in regard to the birth, and passion, and resurrection which took place in the time of the government of Pontius Pilate, being truly and certainly accomplished by Jesus Christ, who is our hope,\footnote{1 Tim. i. 1.} from which may no one of you ever be turned aside.

\textsc{Longer.}~These things [I address to you], my beloved, not that I know any of you to be in such a state;\footnote{i.e., addicted to the error of Judaizing.} but, as less than any of you, I desire to guard you beforehand, that ye fall not upon the hooks of vain doctrine, but that you may rather attain to a full assurance in Christ, who was begotten by the Father before all ages, but was afterwards born of the Virgin Mary without any intercourse with man. He also lived a holy life, and healed every kind of sickness and disease among the people, and wrought signs and wonders for the benefit of men; and to those who had fallen into the error of polytheism He made known the one and only true God, His Father, and underwent the passion, and endured the cross at the hands of the Christ-killing Jews, under Pontius Pilate the governor and Herod the king. He also died, and rose again, and ascended into the heavens to Him that sent Him, and is sat down at His right hand, and shall come at the end of the world, with His Father’s glory, to judge the living and the dead, and to render to every one according to his works.\footnote{2 Tim. iv. 1; Rom. ii. 6.} He who knows these things with a full assurance, and believes them, is happy; even as ye are now the lovers of God and of Christ, in the full assurance of our hope, from which may no one of us\footnote{Some read, “of you.”} ever be turned aside!

\xsection{Chapter XII.—Ye are superior to me.}

\textsc{Shorter.}~May I enjoy you in all respects, if indeed I be worthy! For though I am bound, I am not worthy to be compared to any of you that are at liberty. I know that ye are not puffed up, for ye have Jesus Christ in yourselves. And all the more when I commend you, I know that ye cherish modesty\footnote{Literally, “are reverent.”} of spirit; as it is written, “The righteous man is his own accuser.”\footnote{Prov. xviii. 17. (LXX).}

\textsc{Longer.}~May I enjoy you in all respects, if indeed I be worthy! For though I am bound, I am not worthy to be compared to one of you that are at liberty. I know that ye are not puffed up, for ye have Jesus in yourselves. And all the more when I commend you, I know that ye cherish modesty\footnote{Literally, “are reverent.”} of spirit; as it is written, “The righteous man is his own accuser;”\footnote{Prov. xviii. 17. (LXX).} and again, “Declare thou first thine iniquities, that thou mayest be justified;”\footnote{Isa. xliii. 26.} and again, “When ye shall have done all things that are commanded you, say, We are unprofitable servants;”\footnote{Luke xvii. 10.} “for that which is highly esteemed among men is abomination in the sight of God.”\footnote{Luke xvi. 15.} For says [the Scripture], “God be merciful to me a sinner.”\footnote{Luke xviii. 13.} Therefore those great ones, Abraham and Job,\footnote{Some read, “Jacob.”} styled themselves “dust and ashes” \footnote{Gen. xviii. 27; Job xxx. 19.} before God. And David says, “Who am I before Thee, O Lord, that Thou hast glorified me hitherto?”\footnote{1 Chron. xvii. 16.} And Moses, who was “the meekest of all men,”\footnote{Num. xii. 3.} saith to God, “I am of a feeble voice, and of a slow tongue.”\footnote{Ex. iv. 10.} Be ye therefore also of a humble spirit, that ye may be exalted; for “he that abaseth himself shall be exalted, and he that exalteth himself shall be abased.”\footnote{Luke xiv. 11.}

\xsection{Chapter XIII.—Be established in faith and unity.}

\textsc{Shorter.}~Study, therefore, to be established in the doctrines of the Lord and the apostles, that so all things, whatsoever ye do, may prosper both in the flesh and spirit; in faith and love; in the Son, and in the Father, and in the Spirit; in the beginning and in the end; with your most admirable bishop, and the well-compacted spiritual crown of your presbytery, and the deacons who are according to God. Be ye subject to the bishop, and to one another, as Jesus Christ to the Father,  according to the flesh, and the apostles to Christ, and to the Father, and to the Spirit; that so there may be a union both fleshly and spiritual.

\textsc{Longer.}~Study, therefore, to be established in the doctrines of the Lord and the apostles, that so all things, whatsoever ye do, may prosper, both in the flesh and spirit, in faith and love, with your most admirable bishop, and the well-compacted\footnote{Literally, “well-woven.”} spiritual crown of your presbytery, and the deacons who are according to God. Be ye subject to the bishop, and to one another, as Christ to the Father, that there may be a unity according to God among you.

\xsection{Chapter XIV.—Your prayers requested.}

\textsc{Shorter.}~Knowing as I do that ye are full of God, I have but briefly exhorted you. Be mindful of me in your prayers, that I may attain to God; and of the Church which is in Syria, whence I am not worthy to derive my name: for I stand in need of your united prayer in God, and your love, that the Church which is in Syria may be deemed worthy of being refreshed\footnote{Literally, “of being sprinkled with dew.”} by your Church.

\textsc{Longer.}~Knowing as I do that ye are full of all good, I have but briefly exhorted you in the love of Jesus Christ. Be mindful of me in your prayers, that I may attain to God; and of the Church which is in Syria, of whom I am not worthy to be called bishop. For I stand in need of your united prayer in God, and of your love, that the Church which is in Syria may be deemed worthy, by your good order, of being edified\footnote{Literally, “of being fed as by a shepherd.”} in Christ.

\xsection{Chapter XV.—Salutations.}

\textsc{Shorter.}~The Ephesians from Smyrna (whence I also write to you), who are here for the glory of God, as ye also are, who have in all things refreshed me, salute you, along with Polycarp, the bishop of the Smyrnæans. The rest of the Churches, in honour of Jesus Christ, also salute you. Fare ye well in the harmony of God, ye who have obtained the inseparable Spirit, who is Jesus Christ.

\textsc{Longer.}~The Ephesians from Smyrna (whence I also write to you), who are here for the glory of God, as ye also are, who have in all things refreshed me, salute you, as does also Polycarp. The rest of the Churches, in honour of Jesus Christ, also salute you. Fare ye well in harmony, ye who have obtained the inseparable Spirit, in Christ Jesus, by the will of God.

\xchapter{Epistle to the Trallians}

\textsc{Shorter.}~\emph{Ignatius, who is also called Theophorus, to the holy Church which is at Tralles, in Asia, beloved of God, the Father of Jesus Christ, elect, and worthy of God, possessing peace through the flesh, and blood, and passion of Jesus Christ, who is our hope, through our rising again to Him,}\footnote{Some render, “in the resurrection which is by Him.”}\emph{ which also I salute in its fulness,}\footnote{Either, “the whole members of the Church,” or, “in the fulness of blessing.”}\emph{ and in the apostolical character,}\footnote{Either, “as an apostle,” or, “in the apostolic form.”}\emph{ and wish abundance of happiness.}

\textsc{Longer.}~\emph{Ignatius, who is also called Theophorus, to the holy Church which is at Tralles, beloved by God the Father, and Jesus Christ, elect, and worthy of God, possessing peace through the flesh and Spirit of Jesus Christ, who is our hope, in His passion by the cross and death, and in His resurrection, which also I salute in its fulness,}\footnote{Either, “the whole members of the Church,” or, “in the fulness of blessing.”}\emph{ and in the apostolical character,}\footnote{Either, “as an apostle,” or, “in the apostolic form.”}\emph{ and wish abundance of happiness.}

\xsection{Chapter I.—Acknowledgment of their excellence.}

\textsc{Shorter.}~\textsc{I know} that ye possess an unblameable and sincere mind in patience, and that not only in present practice,\footnote{Literally, “not according to use, but according to nature.”} but according to inherent nature, as Polybius your bishop has shown me, who has come to Smyrna by the will of God and Jesus Christ, and so sympathized in the joy which I, who am bound in Christ Jesus, possess, that I beheld your whole multitude in him. Having therefore received through him the testimony of your good-will, according to God, I gloried to find you, as I knew you were, the followers of God.

\textsc{Longer.}~\textsc{I know} that ye possess an unblameable and sincere mind in patience, and that not only for present use,\footnote{Literally, “not for use, but for a possession.”} but as a permanent possession, as Polybius your bishop has shown me, who has come to Smyrna by the will of God the Father, and the Lord Jesus Christ, His Son, with the co-operation of the Spirit, and so sympathized in the joy which I, who am bound in Christ Jesus, possess, that I beheld your whole multitude in Him. Having therefore received through him the testimony of your good-will according to God, I gloried to find that you were the followers of Jesus Christ the Saviour.

\xsection{Chapter II.—Be subject to the bishop, etc.}

\textsc{Shorter.}~For, since ye are subject to the bishop as to Jesus Christ, ye appear to me to live not after the manner of men, but according to Jesus Christ, who died for us, in order, by believing in His death, ye may escape from death. It is therefore necessary that, as ye indeed do, so without the bishop ye should do nothing, but should also  be subject to the presbytery, as to the apostle of Jesus Christ, who is our hope, in whom, if we live, we shall [at last] be found. It is fitting also that the deacons, as being [the ministers] of the mysteries of Jesus Christ, should in every respect be pleasing to all.\footnote{It is doubtful whether this exhortation is addressed to the deacons or people; whether the former are urged in all respects to please the latter, or the latter in all points to be pleased with the former.} For they are not ministers of meat and drink, but servants of the Church of God. They are bound, therefore, to avoid all grounds of accusation [against them], as they would do fire.

\textsc{Longer.}~Be ye subject to the bishop as to the Lord, for “he watches for your souls, as one that shall give account to God.”\footnote{Heb. xiii. 17.} Wherefore also, ye appear to me to live not after the manner of men, but according to Jesus Christ, who died for us, in order that, by believing in His death, ye may by baptism be made partakers of His resurrection. It is therefore necessary, whatsoever things ye do, to do nothing without the bishop. And be ye subject also to the presbytery, as to the apostles of Jesus Christ, who is our hope, in whom, if we live, we shall be found in Him. It behoves you also, in every way, to please the deacons, who are [ministers] of the mysteries of Christ Jesus; for they are not ministers of meat and drink, but servants of the Church of God. They are bound, therefore, to avoid all grounds of accusation [against them], as they would a burning fire. Let them, then, prove themselves to be such.

\xsection{Chapter III.—Honour the deacons, etc.}

\textsc{Shorter.}~In like manner, let all reverence the deacons as an appointment\footnote{Literally, “commandment.” The text, which is faulty in the \textsc{ms.}, has been amended as above by Smith.} of Jesus Christ, and the bishop as Jesus Christ, who is the Son of the Father, and the presbyters as the sanhedrim of God, and assembly of the apostles. Apart from these, there is no Church.\footnote{Literally, “no Church is called.”} Concerning all this, I am persuaded that ye are of the same opinion. For I have received the manifestation\footnote{Or, “pattern.”} of your love, and still have it with me, in your bishop, whose very appearance is highly instructive,\footnote{Literally, “great instruction.”} and his meekness of itself a power; whom I imagine even the ungodly must reverence, seeing they are\footnote{Some here follow a text similar to that of the longer recension.} also pleased that I do not spare myself. But shall I, when permitted to write on this point, reach such a height of self-esteem, that though being a condemned\footnote{Both the text and meaning are here very doubtful; some follow the reading of the longer recension.} man, I should issue commands to you as if I were an apostle?

\textsc{Longer.}~And do ye reverence them as Christ Jesus, of whose place they are the keepers, even as the bishop is the representative of the Father of all things, and the presbyters are the sanhedrim of God, and assembly\footnote{Or, “conjunction.”} of the apostles of Christ. Apart from these there is no elect Church, no congregation of holy ones, no assembly of saints. I am persuaded that ye also are of this opinion. For I have received the manifestation\footnote{Or, “pattern.”} of your love, and still have it with me, in your bishop, whose very appearance is highly instructive, and his meekness of itself a power; whom I imagine even the ungodly must reverence. Loving you as I do, I avoid writing in any severer strain to you, that I may not seem harsh to any, or wanting [in tenderness]. I am indeed bound for the sake of Christ, but I am not yet worthy of Christ. But when I am perfected, perhaps I shall then become so. I do not issue orders like an apostle.

\xsection{Chapter IV.—I have need of humility.}

\textsc{Shorter.}~I have great knowledge in God,\footnote{Literally, “I know many things in God.”} but I restrain myself, lest, I should perish through boasting. For now it is needful for me to be the more fearful; and not give heed to those that puff me up. For they that speak to me [in the way of commendation] scourge me. For I do indeed desire to suffer, but I know not if I be worthy to do so. For this longing, though it is not manifest to many, all the more vehemently assails me.\footnote{A different turn altogether is given to this passage in the longer recension.} I therefore have need of meekness, by which the prince of this world is brought to nought.

\textsc{Longer.}~But I measure myself, that I may not perish through boasting: but it is good to glory in the Lord.\footnote{1 Cor. i. 31.} And even though I were established\footnote{Or, “confirmed.”} in things pertaining to God, yet then would it befit me to be the more fearful, and not give heed to those that vainly puff me up. For those that commend me scourge me. [I do indeed desire to suffer\footnote{Omitted in the \textsc{ms.}}], but I know not if I be worthy to do so. For the envy of the wicked one is not visible to many, but it wars against me. I therefore have need of meekness, by which the devil, the prince of this world, is brought to nought.

\xsection{Chapter V.—I will not teach you profound doctrines.}

\textsc{Shorter.}~Am I not able to write to you of heavenly things? But I fear to do so, lest I should inflict injury on you who are but babes [in Christ]. Pardon me in this respect, lest, as not being able to receive [such doctrines], ye should be strangled by them. For even I, though I am bound [for Christ], yet am not on that account able to understand heavenly things, and the places\footnote{Or, “stations.”} of the angels, and their gatherings under their respective princes, things visible and invisible. Without reference to such abstruse subjects, I am still but a learner [in other respects\footnote{Literally, “passing by this;” but both text and meaning are very doubtful.}]; for many things are wanting to us, that we come not short of God.

\textsc{Longer.}~For might\footnote{\textgreek{ἐβουλόμην} apparently by mistake for \textgreek{ἐδυνάμην}.} not I write to you things more full of mystery? But I fear to do so, lest I should inflict injury on you who are but babes [in Christ]. Pardon me in this respect, lest, as not being able to receive their weighty import,\footnote{Literally, “their force.”} ye should be strangled by them. For even I, though I am bound [for Christ], and am able to understand heavenly things, the angelic orders, and the different sorts\footnote{Or, “varieties of.”} of angels and hosts, the distinctions between powers and dominions, and the diversities between thrones and authorities, the mightiness of the Æons, and the pre-eminence of the cherubim and seraphim, the sublimity of the spirit, the kingdom of the Lord, and above all, the incomparable majesty of Almighty God—though I am acquainted with these things, yet am I not therefore by any means perfect; nor am I such a disciple as Paul or Peter. For many things are yet wanting to me, that I may not fall short of God.

\xsection{Chapter VI.—Abstain from the poison of heretics.}

\textsc{Shorter.}~I therefore, yet not I, but the love of Jesus Christ, entreat you that ye use Christian nourishment only, and abstain from herbage of a different kind; I mean heresy. For those\footnote{The ellipsis in the original is here very variously supplied.} [that are given to this] mix\footnote{Literally, “interweave.”} up Jesus Christ with their own poison, speaking things which are unworthy of credit, like those who administer a deadly drug in sweet wine, which he who is ignorant of does greedily\footnote{Or, “sweetly.”} take, with a fatal pleasure\footnote{The construction is here difficult and doubtful.} leading to his own death.

\textsc{Longer.}~I therefore, yet not I, out the love of Jesus Christ, “entreat you that ye all speak the same thing, and that there be no divisions among you; but that ye be perfectly joined together in the same mind, and in the same judgment.”\footnote{1 Cor. i. 10.} For there are some vain talkers\footnote{Tit. i. 10.} and deceivers, not Christians, but Christ-betrayers,\footnote{Literally, “Christ-sellers.”} bearing about the name of Christ in deceit, and “corrupting the word”\footnote{2 Cor. ii. 17.} of the Gospel; while they intermix the poison of their deceit with their persuasive talk,\footnote{Literally, “sweet address.”} as if they mingled aconite with sweet wine, that so he who drinks, being deceived in his taste by the very great sweetness of the draught, may incautiously meet with his death. One of the ancients gives us this advice, “Let no man be called good who mixes good with evil.”\footnote{\emph{Apost. Constitutions}, vi. 13.} For they speak of Christ, not that they may preach Christ, but that they may reject Christ; and they speak\footnote{Supplied from the old Latin version.} of the law, not that they may establish the law, but that they may proclaim things contrary to it. For they alienate Christ from the Father, and the law from Christ. They also calumniate His being born of the Virgin; they are ashamed of His cross; they deny His passion; and they do not believe His resurrection. They introduce God as a Being unknown; they suppose Christ to be unbegotten; and as to the Spirit, they do not admit that He exists. Some of them say that the Son is a mere man, and that the Father, Son, and Holy Spirit are but the same person, and that the creation is the work of God, not by Christ, but by some other strange power.

\xsection{Chapter VII.—The same continued.}

\textsc{Shorter.}~Be on your guard, therefore, against such persons. And this will be the case with you if you are not puffed up, and continue in intimate union with\footnote{Literally, “unseparated from.”} Jesus Christ our God, and the  bishop, and the enactments of the apostles. He that is within the altar is pure, but\footnote{This clause is inserted from the ancient Latin version.} he that is without is not pure; that is, he who does anything apart from the bishop, and presbytery, and deacons,\footnote{The text has “deacon.”} such a man is not pure in his conscience.

\textsc{Longer.}~Be on your guard, therefore, against such persons, that ye admit not of a snare for your own souls. And act so that your life shall be without offence to all men, lest ye become as “a snare upon a watch-tower, and as a net which is spread out.”\footnote{Hos. v. 1.} For “he that does not heal himself in his own works, is the brother of him that destroys himself.”\footnote{Prov. xviii. 9 (LXX).} If, therefore, ye also put away conceit, arrogance, disdain, and haughtiness, it will be your privilege to be inseparably united to God, for “He is nigh unto those that fear Him.”\footnote{Ps. lxxxv. 9.} And says He, “Upon whom will I look, but upon him that is humble and quiet, and that trembles at my words?”\footnote{Isa. lxvi. 2.} And do ye also reverence your bishop as Christ Himself, according as the blessed apostles have enjoined you. He that is within the altar is pure, wherefore also he is obedient to the bishop and presbyters: but he that is without is one that does anything apart from the bishop, the presbyters, and the deacons. Such a person is defiled in his conscience, and is worse than an infidel. For what is the bishop but one who beyond all others possesses all power and authority, so far as it is possible for a man to possess it, who according to his ability has been made an imitator of the Christ of God?\footnote{Some render, “being a resemblance according to the power of Christ.”} And what is the presbytery but a sacred assembly, the counsellors and assessors of the bishop? And what are the deacons but imitators of the angelic powers,\footnote{Some read, “imitators of Christ, ministering to the bishop, as Christ to the Father.”} fulfilling a pure and blameless ministry unto him, as the holy Stephen did to the blessed James, Timothy and Linus to Paul, Anencletus and Clement to Peter? He, therefore, that will not yield obedience to such, must needs be one utterly without God, an impious man who despises Christ, and depreciates His appointments.

\xsection{Chapter VIII.—Be on your guard against the snares of the devil.}

\textsc{Shorter.}~Not that I know there is anything of this kind among you; but I put you on your guard, inasmuch as I love you greatly, and foresee the snares of the devil. Wherefore, clothing\footnote{Literally, “taking up.”} yourselves with meekness, be ye renewed\footnote{Or, “renew yourselves.”} in faith, that is the flesh of the Lord, and in love, that is the blood of Jesus Christ. Let no one of you cherish any grudge against his neighbour. Give no occasion to the Gentiles, lest by means of a few foolish men the whole multitude [of those that believe] in God be evil spoken of. For, “Woe to him by whose vanity my name is blasphemed among any.”\footnote{Isa. lii. 5.}

\textsc{Longer.}~Now I write these things unto you, not that I know there are any such persons among you; nay, indeed I hope that God will never permit any such report to reach my ears, He “who spared not His Son for the sake of His holy Church.”\footnote{Rom. viii. 32.}  But foreseeing the snares of the wicked one, I arm you beforehand by my admonitions, as my beloved and faithful children in Christ, furnishing you with the means of protection\footnote{Literally, “making you drink beforehand what will preserve you.”} against the deadly disease of unruly men, by which do ye flee from the disease\footnote{Or, “from which disease.”} [referred to] by the good-will of Christ our Lord. Do ye therefore, clothing\footnote{Literally, “taking up.”} yourselves with meekness, become the imitators of His sufferings, and of His love, wherewith\footnote{Comp. Eph. ii. 4.} He loved us when He gave Himself a ransom\footnote{Comp. 1 Tim. ii. 6.} for us, that He might cleanse us by His blood from our old ungodliness, and bestow life on us when we were almost on the point of perishing through the depravity that was in us. Let no one of you, therefore, cherish any grudge against his neighbour. For says our Lord, “Forgive, and it shall be forgiven unto you.”\footnote{Matt. vi. 14.} Give no occasion to the Gentiles, lest “by means of a few foolish men the word and doctrine [of Christ] be blasphemed.”\footnote{1 Tim. vi. 1; Tit. ii. 5.} For says the prophet, as in the person of God, “Woe to him by whom my name is blasphemed among the Gentiles.”\footnote{Isa. lii. 5.}

\xsection{Chapter IX.—Reference to the history of Christ.}

\textsc{Shorter.}~Stop your ears, therefore, when any one speaks to you at variance with\footnote{Literally, “apart from.”}  Jesus Christ, who was descended from David, and was also of Mary; who was truly born, and did eat and drink. He was truly persecuted under Pontius Pilate; He was truly crucified, and [truly] died, in the sight of beings in heaven, and on earth, and under the earth. He was also truly raised from the dead, His Father quickening Him, even as after the same manner His Father will so raise up us who believe in Him by Christ Jesus, apart from whom we do not possess the true life.

\textsc{Longer.}~Stop your ears, therefore, when any one speaks to you at variance with\footnote{Literally, “apart from.”} Jesus Christ, the Son of God, who was descended from David, and was also of Mary; who was truly begotten of God and of the Virgin, but not after the same manner. For indeed God and man are not the same. He truly assumed a body; for “the Word was made flesh,”\footnote{John i. 14.} and lived upon earth without sin. For says He, “Which of you convicteth me of sin?”\footnote{John viii. 46.} He did in reality both eat and drink. He was crucified and died under Pontius Pilate. He really, and not merely in appearance, was crucified, and died, in the sight of beings in heaven, and on earth, and under the earth. By those in heaven I mean such as are possessed of incorporeal natures; by those on earth, the Jews and Romans, and such persons as were present at that time when the Lord was crucified; and by those under the earth, the multitude that arose along with the Lord. For says the Scripture, “Many bodies of the saints that slept arose,”\footnote{Matt. xxvii. 52.} their graves being opened. He descended, indeed, into Hades alone, but He arose accompanied by a multitude; and rent asunder that means\footnote{Literally, “hedge,” or “fence.”} of separation which had existed from the beginning of the world, and cast down its partition-wall. He also rose again in three days, the Father raising Him up; and after spending forty days with the apostles, He was received up to the Father, and “sat down at His right hand, expecting till His enemies are placed under His feet.”\footnote{Heb. x. 12, 13.} On the day of the preparation, then, at the third hour, He received the sentence from Pilate, the Father permitting that to happen; at the sixth hour He was crucified; at the ninth hour He gave up the ghost; and before sunset He was buried.\footnote{Some read, “He was taken down from the cross, and laid in a new tomb.”} During the Sabbath He continued under the earth in the tomb in which Joseph of Arimathæa had laid Him. At the dawning of the Lord’s day He arose from the dead, according to what was spoken by Himself, “As Jonah was three days and three nights in the whale’s belly, so shall the Son of man also be three days and three nights in the heart of the earth.”\footnote{Matt. xii. 40.} The day of the preparation, then, comprises the passion; the Sabbath embraces the burial; the Lord’s Day contains the resurrection.

\xsection{Chapter X.—The reality of Christ’s passion.}

\textsc{Shorter.}~But if, as some that are without God, that is, the unbelieving, say, that He only seemed to suffer (they themselves only seeming to exist), then why am I in bonds? Why do I long to be exposed to\footnote{Literally, “to fight with.”} the wild beasts? Do I therefore die in vain?\footnote{Some read this and the following clause affirmatively, instead of interrogatively.} Am I not then guilty of falsehood\footnote{The meaning is, that is they spoke the truth concerning the phantasmal character of Christ’s death, then Ignatius was guilty of a practical falsehood in suffering for what was false.} against [the cross of] the Lord?

\textsc{Longer.}~But if, as some that are without God, that is, the unbelieving, say, He became man in appearance [only], that He did not in reality take unto Him a body, that He died in appearance [merely], and did not in very deed suffer, then for what reason am I now in bonds, and long to be exposed to\footnote{Literally, “to fight with.”} the wild beasts? In such a case, I die in vain, and am guilty of falsehood\footnote{The meaning is, that if they spoke the truth concerning the phantasmal character of Christ’s death, then Ignatius was guilty of a practical falsehood in suffering for what was false.} against the cross of the Lord. Then also does the prophet in vain declare, “They shall look on Him whom they have pierced, and mourn over themselves as over one beloved.”\footnote{Zech. xii. 10.} These men, therefore, are not less unbelievers than were those that crucified Him. But as for me, I do not place my hopes in one who died for me in appearance, but in reality. For that which is false is quite abhorrent to the truth. Mary then did truly conceive a body which had God inhabiting it. And God the Word was truly born of the Virgin, having clothed Himself with a body of like passions with our own. He who forms all men in the womb, was Himself really in the womb, and made for Himself a body of the seed of the Virgin, but without any intercourse of man. He was carried in the womb, even as we are, for the usual period of time; and was really born, as we also are; and was in reality nourished with milk, and partook of common meat and drink, even as we do. And when He had lived among men for thirty years, He was baptized by John, really and not in appearance; and when He had preached the Gospel three years, and done signs and wonders, He who was Himself the Judge was judged by the Jews, falsely so called, and by Pilate the governor; was scourged, was smitten on the cheek, was spit upon; He wore a crown of thorns and a purple robe; He was condemned: He was crucified in reality, and not in appearance, not in imagination, not in deceit. He really died, and was buried, and rose from the dead, even as He prayed in a certain place, saying, “But do Thou, O Lord, raise me up again, and I shall recompense them.”\footnote{Ps. xli. 10.} And the Father, who always hears Him,\footnote{Comp. John xi. 42.} answered and said, “Arise, O God, and judge the earth; for Thou shall receive all the heathen for Thine inheritance.”\footnote{Ps. lxxxii. 8.} The Father, therefore, who raised Him up, will also raise us up through Him, apart from whom no one will attain to true life. For says He, “I am the life; he that believeth in me, even though he die, shall live: and every one that liveth and believeth in me, even though he die, shall live for ever.”\footnote{John xi. 25, 26.} Do ye therefore flee from these ungodly heresies; for they are the inventions of the devil, that serpent who was the author of evil, and who by means of the woman deceived Adam, the father of our race.

\xsection{Chapter XI.—Avoid the deadly errors of the Docetæ.}

\textsc{Shorter.}~Flee, therefore, those evil offshoots [of Satan], which produce death-bearing fruit, whereof if any one tastes, he instantly dies. For these men are not the planting of the Father. For if they were, they would appear as branches of the cross, and their fruit would be incorruptible. By it\footnote{i.e., the cross.} He calls you through His passion, as being His members. The head, therefore, cannot be born by itself, without its members; God, who is [the Saviour] Himself, having promised their union.\footnote{Both text and meaning here are doubtful.}

\textsc{Longer.}~Do ye also avoid those wicked offshoots of his,\footnote{i.e., Satan’s.} Simon his firstborn son, and Menander, and Basilides, and all his wicked mob of followers,\footnote{Literally, “loud, confused noise.”} the worshippers of a man, whom also the prophet Jeremiah pronounces accursed.\footnote{The Ebionites, who denied the divine nature of our Lord, are here referred to.} Flee also the impure Nicolaitanes, falsely so called,\footnote{It seems to be here denied that Nicolas was the founder of this school of heretics.} who are lovers of pleasure, and given to calumnious speeches. Avoid also the children of the evil one, Theodotus and Cleobulus, who produce death-bearing fruit, whereof if any one tastes, he instantly dies, and that not a mere temporary death, but one that shall endure for ever. These men are not the planting of the Father, but are an accursed brood. And says the Lord, “Let every plant which my heavenly Father has not planted be rooted up.”\footnote{Matt. xv. 13.} For if they had been branches of the Father, they would not have been “enemies of the cross of Christ,”\footnote{Phil. iii. 18.} but rather of those who “killed the Lord of glory.”\footnote{1 Cor. ii. 8.} But now, by denying the cross, and being ashamed of the passion, they cover the transgression of the Jews, those fighters against God, those murderers of the Lord; for it were too little to style them merely murderers of the prophets. But Christ invites you to [share in] His immortality, by His passion and resurrection, inasmuch as ye are His members.

\xsection{Chapter XII.—Continue in unity and love.}

\textsc{Shorter.}~I salute you from Smyrna, together with the Churches of God which are with me, who have refreshed me in all things, both in the flesh and in the spirit. My bonds, which I carry about with me for the sake of Jesus Christ (praying that I may attain to God), exhort you. Continue in harmony among yourselves, and in prayer with one another; for it becomes every one of you, and especially the presbyters, to refresh the bishop, to the honour of the Father, of Jesus Christ, and of the apostles. I entreat you in love to hear me, that I may not, by having written, be a testimony against you. And do ye also pray for me, who have need of your love, along with the mercy of God, that I may be worthy of the lot for which I am destined, and that I may not be found reprobate.

\textsc{Longer.}~I salute you from Smyrna, together with the Churches of God which are with me, whose rulers have refreshed me in every respect, both in the flesh and in the spirit. My bonds, which I carry about with me for the sake of Jesus Christ (praying that I may attain to God), exhort you. Continue in harmony among yourselves, and in supplication; for it becomes every one of you, and especially the presbyters, to refresh the bishop, to the honour of the Father, and to the honour of Jesus Christ and of the apostles. I entreat you in love to hear me, that I may not, by having thus written, be a testimony against you. And do ye also pray for me, who have need of your love, along with the mercy of God, that I may be thought worthy to attain the lot for which I am now designed, and that I may not be found reprobate.

\xsection{Chapter XIII.—Conclusion.}

\textsc{Shorter.}~The love of the Smyrnæans and Ephesians salutes you. Remember in your prayers the Church which is in Syria, from which also I am not worthy to receive my appellation, being the last\footnote{i.e., the least.} of them. Fare ye well in Jesus Christ, while ye continue subject to the bishop, as to the command [of God], and in like manner to the presbytery. And do ye, every man, love one another with an undivided heart. Let my spirit be sanctified\footnote{The shorter recension reads \textgreek{ἁγνίζετε}, and the longer also hesitates between this and \textgreek{ἀσπάζεται}. With the former reading the meaning is very obscure: it has been corrected as above to \textgreek{ἁγνίζηται}.} by yours, not only now, but also when I shall attain to God. For I am as yet exposed to danger. But the Father is faithful in Jesus Christ to fulfil both mine and your petitions: in whom may ye be found unblameable.

\textsc{Longer.}~The love of the Smyrnæans and Ephesians salutes you. Remember our Church which is in Syria, from which I am not worthy to receive my appellation, being the last\footnote{i.e., the least.} of those of that place. Fare ye well in the Lord Jesus Christ, while ye continue subject to the bishop, and in like manner to the presbyters and to the deacons. And do ye, every man, love one another with an undivided heart. My spirit salutes you,\footnote{The shorter recension reads \textgreek{ἁγνίζετε}, and the longer also hesitates between this and \textgreek{ἀσπάζεται}. With the former reading the meaning is very obscure: it has been corrected as above to \textgreek{ἁγνίζηται}.} not only now, but also when I shall have attained to God; for I am as yet exposed to danger. But the Father of Jesus Christ is faithful to fulfil both mine and your petitions: in whom may we be found without spot. May I have joy of you in the Lord.

\xchapter{Epistle to the Romans}

\textsc{Shorter.}~\emph{Ignatius, who is also called Theophorus, to the Church which has obtained mercy, through the majesty of the Most High Father, and Jesus Christ, His only-begotten Son; the Church which is beloved and enlightened by the will of Him that willeth all things which are according to the love of Jesus Christ our God, which also presides in the place of the region of the Romans, worthy of God, worthy of honour, worthy of the highest happiness, worthy of praise, worthy of obtaining her every desire, worthy of being deemed holy,}\footnote{Or, “most holy.”}\emph{ and which presides over love, is named from Christ, and from the Father, which I also salute in the name of Jesus Christ, the Son of the Father: to those who are united, both according to the flesh and spirit, to every one of His commandments; who are filled inseparably with the grace of God, and are purified from every strange taint, [I wish] abundance of happiness unblameably, in Jesus Christ our God.}

\textsc{Longer.}~\emph{Ignatius, who is also called Theophorus, to the Church which has obtained mercy, through the majesty of the Most High God the Father, and of Jesus Christ, His only-begotten Son; the Church which is sanctified and enlightened by the will of God, who formed all things that are according to the faith and love of Jesus Christ, our God and Saviour; the Church which presides in the place of the region of the Romans, and which is worthy of God, worthy of honour, worthy of the highest happiness, worthy of praise, worthy of credit,}\footnote{Or as in the shorter recension.}\emph{ worthy of being deemed holy,}\footnote{Or, “most holy.”}\emph{ and which presides over love, is named from Christ, and from the Father, and is possessed of the Spirit, which I also salute in the name of Almighty God, and of Jesus Christ His Son: to those who are united, both according to the flesh and spirit, to every one of His commandments, who are filled inseparably with all the grace of God, and are purified from every strange taint, [I wish] abundance of happiness unblameably, in God, even the Father, and our Lord Jesus Christ.}

\xsection{Chapter I.—As a prisoner, I hope to see you.}

\textsc{Shorter.}~\textsc{Through} prayer\footnote{Some read, “since I have,” leaving out the following “for,” and finding the apodosis in “I hope to salute you.”} to God I have obtained the privilege of seeing your most worthy faces,\footnote{Literally, “worthy of God.”} and have even\footnote{Some read, “which I much desired to do.”} been granted more than I requested; for I hope as a prisoner in Christ Jesus to salute you, if indeed it be the will of God that I be thought worthy of attaining unto the end. For the beginning has been well ordered,  if I may obtain grace to cling to\footnote{Literally, “to receive.”} my lot without hindrance unto the end. For I am afraid of your love,\footnote{He probably refers here, and in what follows, to the influence which their earnest prayers in his behalf might have with God.} lest it should do me an injury. For it is easy for you to accomplish what you please; but it is difficult for me to attain to God, if ye spare me.

\textsc{Longer.}~\textsc{Through} prayer to God I have obtained the privilege of seeing your most worthy faces,\footnote{Literally, “worthy of God.”} even as I earnestly begged might be granted me; for as a prisoner in Christ Jesus I hope to salute you, if indeed it be the will [of God] that I be thought worthy of attaining unto the end. For the beginning has been well ordered, if I may obtain grace to cling to\footnote{Literally, “to receive.”} my lot without hindrance unto the end. For I am afraid of your love,\footnote{He probably refers here, and in what follows, to the influence which their earnest prayers in his behalf might have with God.} lest it should do me an injury. For it is easy for you to accomplish what you please; but it is difficult for me to attain to God, if ye do not spare me,\footnote{Some read \textgreek{γε} instead of \textgreek{μή}, and translate as in shorter recension.} under the pretence of carnal affection.

\xsection{Chapter II.—Do not save me from martyrdom.}

\textsc{Shorter.}~For it is not my desire to act towards you as a man-pleaser,\footnote{Some translate as in longer recension, but there is in the one case \textgreek{ὑμῖν}, and in the other \textgreek{ὑμᾶς}.} but as pleasing God, even as also ye please Him. For neither shall I ever have such [another] opportunity of attaining to God; nor will ye, if ye shall now be silent, ever be entitled to\footnote{Literally, “have to be inscribed to.”} the honour of a better work. For if ye are silent concerning me, I shall become God’s; but if you show your love to my flesh, I shall again have to run my race. Pray, then, do not seek to confer any greater favour upon me than that I be sacrificed to God while the altar is still prepared; that, being gathered together in love, ye may sing praise to the Father, through Christ Jesus, that God has deemed me, the bishop of Syria, worthy to be sent for\footnote{Literally, “to be found and sent for.”} from the east unto the west. It is good to set from the world unto God, that I may rise again to Him.

\textsc{Longer.}~For it is not my desire that ye should please men, but God, even as also ye do please Him. For neither shall I ever hereafter have such an opportunity of attaining to God; nor will ye, if ye shall now be silent, ever be entitled to\footnote{Literally, “have to be inscribed to.”} the honour of a better work. For if ye are silent concerning me, I shall become God’s; but if ye show your love to my flesh, I shall again have to run my race. Pray, then, do not seek to confer any greater favour upon me than that I be sacrificed to God, while the altar is still prepared; that, being gathered together in love, ye may sing praise to the Father, through Christ Jesus, that God has deemed me, the bishop of Syria, worthy to be sent for\footnote{Literally, “to be found and sent for.”} from the east unto the west, and to become a martyr\footnote{The text is here in great confusion.} in behalf of His own precious\footnote{Literally, “beautiful.” Some read, “it is good,” etc.} sufferings, so as to pass from the world to God, that I may rise again unto Him.

\xsection{Chapter III.—Pray rather that I may attain to martyrdom.}

\textsc{Shorter.}~Ye have never envied any one; ye have taught others. Now I desire that those things may be confirmed [by your conduct], which in your instructions ye enjoin [on others]. Only request in my behalf both inward and outward strength, that I may not only speak, but [truly] will; and that I may not merely be called a Christian, but really be found to be one. For if I be truly found [a Christian], I may also be called one, and be then deemed faithful, when I shall no longer appear to the world. Nothing visible is eternal.\footnote{Some read, “good.”} “For the things which are seen are temporal, but the things which are not seen are eternal.”\footnote{2 Cor. iv. 18. This quotation is not found in the old Latin version of the shorter recension.} For our God, Jesus Christ, now that He is with\footnote{Or, “in.”} the Father, is all the more revealed [in  His glory]. Christianity is not a thing\footnote{Literally, “work.”} of silence only, but also of [manifest] greatness.

\textsc{Longer.}~Ye have never envied any one; ye have taught others. Now I desire that those things may be confirmed [by your conduct], which in your instructions ye enjoin [on others]. Only request in my behalf both inward and outward strength, that I may not only speak, but [truly] will, so that I may not merely be called a Christian, but really found to be one. For if I be truly found [a Christian], I may also be called one, and be then deemed faithful, when I shall no longer appear to the world. Nothing visible is eternal. “For the things which are seen are temporal, but the things which are not seen are eternal.”\footnote{2 Cor. iv. 18. This quotation is not found in the old Latin version of the shorter recension.} The Christian is not the result\footnote{Literally, “work.”} of persuasion, but of power.\footnote{The meaning is here doubtful.} When he is hated by the world, he is beloved of God. For says [the Scripture], “If ye were of this world, the world would love its own; but now ye are not of the world, but I have chosen you out of it: continue in fellowship with me.”\footnote{John xv. 19.}

\xsection{Chapter IV.—Allow me to fall a prey to the wild beasts.}

\textsc{Shorter.}~I write to the Churches, and impress on them all, that I shall willingly die for God, unless ye hinder me. I beseech of you not to show an unseasonable good-will towards me. Suffer me to become food for the wild beasts, through whose instrumentality it will be granted me to attain to God. I am the wheat of God, and let me be ground by the teeth of the wild beasts, that I may be found the pure bread of Christ. Rather entice the wild beasts, that they may become my tomb, and may leave nothing of my body; so that when I have fallen asleep [in death], I may be no trouble to any one. Then shall I truly be a disciple of Christ, when the world shall not see so much as my body. Entreat Christ for me, that by these instruments\footnote{i.e., by the teeth of the wild beasts.} I may be found a sacrifice [to God]. I do not, as Peter and Paul, issue commandments unto you. They were apostles; I am but a condemned man: they were free,\footnote{“Free,” probably from human infirmity.} while I am, even until now, a servant. But when I suffer, I shall be the freed-man of Jesus, and shall rise again emancipated in Him. And now, being a prisoner, I learn not to desire anything worldly or vain.

\textsc{Longer.}~I write to all the Churches, and impress on them all, that I shall willingly die for God, unless ye hinder me. I beseech of you not to show an unseasonable good-will towards me. Suffer me to become food for the wild beasts, through whose instrumentality it will be granted me to attain to God. I am the wheat of God, and am ground by the teeth of the wild beasts, that I may be found the pure bread of God. Rather entice the wild beasts, that they may become my tomb, and may leave nothing of my body; so that when I have fallen asleep [in death], I may not be found troublesome to any one. Then shall I be a true disciple of Jesus Christ, when the world shall not see so much as my body. Entreat the Lord for me, that by these instruments\footnote{i.e., by the teeth of the wild beasts.} I may be found a sacrifice to God. I do not, as Peter and Paul, issue commandments unto you. They were apostles of Jesus Christ, but I am the very least [of believers]: they were free,\footnote{“Free,” probably from human infirmity.} as the servants of God; while I am, even until now, a servant. But when I suffer, I shall be the freed-man of Jesus Christ, and shall rise again emancipated in Him. And now, being in bonds for Him, I learn not to desire anything worldly or vain.

\xsection{Chapter V.—I desire to die.}

\textsc{Shorter.}~From Syria even unto Rome I fight with beasts,\footnote{Comp. 1 Cor. xv. 32, where the word is also used figuratively.} both by land and sea, both by night and day, being bound to ten leopards, I mean a band of soldiers, who, even when they receive benefits,\footnote{Probably the soldiers received gifts from the Christians, to treat Ignatius with kindness.} show themselves all the worse. But I am the more instructed by their injuries [to act as a disciple of Christ]; “yet am I not thereby justified.”\footnote{1 Cor. iv. 4.} May I enjoy the wild beasts that are prepared for me; and I pray they may be found eager to rush upon me, which also I will entice to devour me speedily, and not deal with me as with some, whom, out of fear, they have not touched. But if they be unwilling to assail me, I will compel them to do so. Pardon me  [in this]: I know what is for my benefit. Now I begin to be a disciple. And let no one, of things visible or invisible, envy\footnote{In the shorter recension there is \textgreek{ζηλώσῃ}, and in the longer \textgreek{ζηλῶσαι}; hence the variety of rendering, but the translation is by no means certain.} me that I should attain to Jesus Christ. Let fire and the cross; let the crowds of wild beasts; let tearings,\footnote{Some deem this and the following word spurious.} breakings, and dislocations of bones; let cutting off of members; let shatterings of the whole body; and let all the dreadful\footnote{Literally, “evil.”} torments of the devil come upon me: only let me attain to Jesus Christ.

\textsc{Longer.}~From Syria even unto Rome I fight with beasts,\footnote{Comp. 1 Cor. xv. 32, where the word is also used figuratively.} both by land and sea, both by night and day, being bound to ten leopards, I mean a band of soldiers, who, even when they receive benefits,\footnote{Probably the soldiers received gifts from the Christians, to treat Ignatius with kindness.} show themselves all the worse. But I am the more instructed by their injuries [to act as a disciple of Christ]; “yet am I not thereby justified.”\footnote{1 Cor. iv. 4.} May I enjoy the wild beasts that are prepared for me; and I pray that they may be found eager to rush upon me, which also I will entice to devour me speedily, and not deal with me as with some, whom, out of fear, they have not touched. But if they be unwilling to assail me, I will compel them to do so. Pardon me [in this] I know what is for my benefit. Now I begin to be a disciple, and have\footnote{In the shorter recension there is \textgreek{ζηλώσῃ}, and in the longer \textgreek{ζηλῶσαι}; hence the variety of rendering, but the translation is by no means certain.} no desire after anything visible or invisible, that I may attain to Jesus Christ. Let fire and the cross; let the crowds of wild beasts; let breakings, tearings, and separations of bones; let cutting off of members; let bruising to pieces of the whole body; and let the very torment of the devil come upon me: only let me attain to Jesus Christ.

\xsection{Chapter VI.—By death I shall attain true life.}

\textsc{Shorter.}~All the pleasures of the world, and all the kingdoms of this earth,\footnote{Literally, “this age.”} shall profit me nothing. It is better for me to die in behalf of\footnote{Literally, “into.”} Jesus Christ, than to reign over all the ends of the earth. “For what shall a man be profited, if he gain the whole world, but lose his own soul?”\footnote{Matt. xvi. 26. Some omit this quotation.} Him I seek, who died for us: Him I desire, who rose again for our sake. This is the gain which is laid up for me. Pardon me, brethren: do not hinder me from living, do not wish to keep me in a state of death;\footnote{Literally, “to die.”} and while I desire to belong to God, do not ye give me over to the world. Suffer me to obtain pure light: when I have gone thither, I shall indeed be a man of God. Permit me to be an imitator of the passion of my God. If any one has Him within himself, let him consider what I desire, and let him have sympathy with me, as knowing how I am straitened.

\textsc{Longer.}~All the ends of the world, and all the kingdoms of this earth,\footnote{Literally, “this age.”} shall profit me nothing. It is better for me to die for the sake of Jesus Christ, than to reign over all the ends of the earth. “For what is a man profited, if he gain the whole world, but lose his own soul?” I long after the Lord, the Son of the true God and Father, even Jesus Christ. Him I seek, who died for us and rose again. Pardon me, brethren: do not hinder me in attaining to life; for Jesus is the life of believers. Do not wish to keep me in a state of death,\footnote{Literally, “to die.”} for life without Christ is death. While I desire to belong to God, do not ye give me over to the world. Suffer me to obtain pure light: when I have gone thither, I shall indeed be a man of God. Permit me to be an imitator of the passion of Christ, my God. If any one has Him within himself, let him consider what I desire, and let him have sympathy with me, as knowing how I am straitened.

\xsection{Chapter VII.—Reason of desiring to die.}

\textsc{Shorter.}~The prince of this world would fain carry me away, and corrupt my disposition towards God. Let none of you, therefore, who are [in Rome] help him; rather be ye on my side, that is, on the side of God. Do not speak of Jesus Christ, and yet set your desires on the world. Let not envy find a dwelling-place among you; nor even should I, when present with you, exhort you to it, be ye persuaded to listen to me, but rather give credit to those things which I now write to you. For though I am alive while I write to you, yet I am eager to die. My love\footnote{Some understand by \emph{love} in this passage, \emph{Christ Himself}; others regard it as referring to \emph{the natural desires of the heart}.} has been crucified, and there is no  fire in me desiring to be fed;\footnote{Literally, “desiring material.”} but there is within me a water that liveth and speaketh,\footnote{The text and meaning are here doubtful. We have followed Hefele, who understands by the water \emph{the Holy Spirit}, and refers to John vii. 38.} saying to me inwardly, Come to the Father. I have no delight in corruptible food, nor in the pleasures of this life. I desire the bread of God, the heavenly bread, the bread of life, which is the flesh of Jesus Christ, the Son of God, who became afterwards of the seed of David and Abraham; and I desire the drink of God, namely His blood, which is incorruptible love and eternal life.

\textsc{Longer.}~The prince of this world would fain carry me away, and corrupt my disposition towards God. Let none of you, therefore, who are [in Rome] help him; rather be ye on my side, that is, on the side of God. Do not speak of Jesus Christ, and yet prefer this world to Him. Let not envy find a dwelling-place among you; nor even should I, when present with you, exhort you to it, be ye persuaded, but rather give credit to those things which I now write to you. For though I am alive while I write to you, yet I am eager to die for the sake of Christ. My love\footnote{Some understand by \emph{love} in this passage, \emph{Christ Himself}; others regard it as referring to \emph{the natural desires of the heart}.} has been crucified, and there is no fire in me that loves anything; but there is living water springing up in me,\footnote{Comp. John iv. 14.} and which says to me inwardly, Come to the Father. I have no delight in corruptible food, nor in the pleasures of this life. I desire the bread of God, the heavenly bread, the bread of life, which is the flesh of Jesus Christ, the Son of God, who became afterwards of the seed of David and Abraham; and I desire the drink, namely His blood, which is incorruptible love and eternal life.

\xsection{Chapter VIII.—Be ye favourable to me.}

\textsc{Shorter.}~I no longer wish to live after the manner of men, and my desire shall be fulfilled if ye consent. Be ye willing, then, that ye also may have your desires fulfilled. I entreat you in this brief letter; do ye give credit to me. Jesus Christ will reveal these things to you, [so that ye shall know] that I speak truly. He\footnote{Some refer this to Ignatius himself.} is the mouth altogether free from falsehood, by which the Father has truly spoken. Pray ye for me, that I may attain [the object of my desire]. I have not written to you according to the flesh, but according to the will of God. If I shall suffer, ye have wished [well] to me; but if I am rejected, ye have hated me.

\textsc{Longer.}~I no longer wish to live after the manner of men, and my desire shall be fulfilled if ye consent. “I am crucified with Christ: nevertheless I live; yet no longer I, since Christ liveth in me.”\footnote{Gal. ii. 20.} I entreat you in this brief letter: do not refuse me; believe me that I love Jesus, who was delivered [to death] for my sake. “What shall I render to the Lord for all His benefits towards me?”\footnote{Ps. cxvi. 12.} Now God, even the Father, and the Lord Jesus Christ, shall reveal these things to you, [so that ye shall know] that I speak truly. And do ye pray along with me, that I may attain my aim in the Holy Spirit. I have not written to you according to the flesh, but according to the will of God. If I shall suffer, ye have loved me; but if I am rejected, ye have hated me.

\xsection{Chapter IX.—Pray for the church in Syria.}

\textsc{Shorter.}~Remember in your prayers the Church in Syria, which now has God for its shepherd, instead of me. Jesus Christ alone will oversee it, and your love [will also regard it]. But as for me, I am ashamed to be counted one of them; for indeed I am not worthy, as being the very last of them, and one born out of due time.\footnote{Comp. 1 Cor. xv. 8, 9.} But I have obtained mercy to be somebody, if I shall attain to God. My spirit salutes you, and the love of the Churches that have received me in the name of Jesus Christ, and not as a mere passer-by. For even those Churches which were not\footnote{Some refer this to the jurisdiction of Ignatius.} near to me in the way, I mean according to the flesh,\footnote{i.e., the outward road he had to travel.} have gone before me,\footnote{Or, “have sent me forward;” comp. Tit. iii. 13.} city by city, [to meet me.]

\textsc{Longer.}~Remember in your prayers the Church which is in Syria, which, instead of me, has now for its shepherd the Lord, who says, “I am the good Shepherd.” And He alone will oversee it, as well as your love towards Him. But as for me, I am ashamed to be counted one of them; for I am not worthy, as being the very last of them, and one born out of due time. But I have obtained mercy to be somebody, if I shall attain to God. My spirit salutes you, and the love of the Churches which have received me in the name of Jesus Christ, and not as a mere passer-by. For even those Churches which were not near to me in the way, have brought me forward, city by city.

\xsection{Chapter X.—Conclusion.}

Now I write these things to you from Smyrna by the Ephesians, who are deservedly most happy. There is also with me, along with many others, Crocus, one dearly beloved by me.\footnote{Literally, “the name desired to me.”} As to those who have gone before me from Syria to Rome for the glory of God, I believe that you are acquainted with them; to whom, [then,] do ye make known that I am at hand. For they are all worthy, both of God and of you; and it is becoming that you should refresh them in all things. I have written these things unto you, on the day before the ninth of the Kalends of September (that\footnote{This clause is evidently an explanatory gloss which has crept into the text.} is, on the twenty-third day of August). Fare ye well to the end, in the patience of Jesus Christ. Amen.

\xchapter{Epistle to the Philadelphians}

\textsc{Shorter.}~\emph{Ignatius, who is also called Theophorus, to the Church of God the Father, and our Lord Jesus Christ, which is at Philadelphia, in Asia, which has obtained mercy, and is established in the harmony of God, and rejoiceth unceasingly}\footnote{Or, “inseparably.”}\emph{ in the passion of our Lord, and is filled with all mercy through his resurrection; which I salute in the blood of Jesus Christ, who is our eternal and enduring joy, especially if [men] are in unity with the bishop, the presbyters, and the deacons, who have been appointed according to the mind of Jesus Christ, whom He has established in security, after His own will, and by His Holy Spirit.}

\textsc{Longer.}~\emph{Ignatius, who is also called Theophorus, to the Church of God the Father, and of the Lord Jesus Christ, which is at Philadelphia, which has obtained mercy through love, and is established in the harmony of God, and rejoiceth unceasingly,}\footnote{Or, “inseparably.”}\emph{ in the passion of our Lord Jesus, and is filled with all mercy through His resurrection; which I salute in the blood of Jesus Christ, who is our eternal and enduring joy, especially to those who are in unity with the bishop, and the presbyters, and the deacons, who have been appointed by the will of God the Father, through the Lord Jesus Christ, who, according to His own will, has firmly established His Church upon a rock, by a spiritual building, not made with hands, against which the winds and the floods have beaten, yet have not been able to overthrow it:}\footnote{Comp. Matt. vii. 25.}\emph{ yea, and may spiritual wickedness never be able to do so, but be thoroughly weakened by the power of Jesus Christ our Lord.}

\xsection{Chapter I.—Praise of the bishop.}

\textsc{Shorter.}~\textsc{Which} bishop,\footnote{The bishop previously referred to.} I know, obtained the ministry which pertains to the common [weal], not of himself, neither by men,\footnote{Comp. Gal. i. 1.} nor through vainglory, but by the love of God the Father, and the Lord Jesus Christ; at whose meekness I am struck with admiration, and who by his silence is able to accomplish more than those who vainly talk. For he is in harmony with the commandments [of God], even as the harp is with its strings. Wherefore my soul declares his mind towards God a happy one, knowing it to be virtuous and perfect, and that his stability as well as freedom from all anger is after the example of the infinite\footnote{Literally, “all.”} meekness of the living God.

\textsc{Longer.}~\textsc{Having} beheld your bishop, I know that he was not selected to undertake the ministry which pertains to the common [weal], either by himself or by men,\footnote{Comp. Gal. i. 1.} or out of vainglory, but by the love of Jesus Christ, and of God the Father, who raised Him from the dead; at whose meekness I am struck with admiration, and who by His silence is able to accomplish more than they who talk a great deal. For he is in harmony with the commandments and ordinances of the Lord, even as the strings are with the harp, and is no less blameless than was Zacharias the priest.\footnote{Luke i. 6.} Wherefore my soul declares his mind towards God a happy one, knowing it to be virtuous and perfect, and that his stability as well as freedom from all anger is after the example of the infinite meekness of the living God.

\xsection{Chapter II.—Maintain union with the bishop.}

\textsc{Shorter.}~Wherefore, as children of light and truth, flee from division and wicked  doctrines; but where the shepherd is, there do ye as sheep follow. For there are many wolves that appear worthy of credit, who, by means of a pernicious pleasure, carry captive\footnote{Comp. 2 Tim. iii. 6.} those that are running towards God; but in your unity they shall have no place.

\textsc{Longer.}~Wherefore, as children of light and truth, avoid the dividing of your unity, and the wicked doctrine of the heretics, from whom “a defiling influence has gone forth into all the earth.”\footnote{Jer. xxiii. 15.} But where the shepherd is, there do ye as sheep follow. For there are many wolves in sheep’s clothing,\footnote{Comp. Matt. vii. 15.} who, by means of a pernicious pleasure, carry captive\footnote{Comp. 2 Tim. iii. 6.} those that are running towards God; but in your unity they shall have no place.

\xsection{Chapter III.—Avoid schismatics.}

\textsc{Shorter.}~Keep yourselves from those evil plants which Jesus Christ does not tend, because they are not the planting of the Father. Not that I have found any division among you, but exceeding purity. For as many as are of God and of Jesus Christ are also with the bishop. And as many as shall, in the exercise of repentance, return into the unity of the Church, these, too, shall belong to God, that they may live according to Jesus Christ. Do not err, my brethren. If any man follows him that makes a schism in the Church, he shall not inherit the kingdom of God. If any one walks according to a strange\footnote{i.e., heretical.} opinion, he agrees not with the passion [of Christ.].

\textsc{Longer.}~Keep yourselves, then, from those evil plants which Jesus Christ does not tend, but that wild beast, the destroyer of men, because they are not the planting of the Father, but the seed of the wicked one. Not that I have found any division among you do I write these things; but I arm you beforehand, as the children of God. For as many as are of Christ are also with the bishop; but as many as fall away from him, and embrace communion with the accursed, these shall be cut off along with them. For they are not Christ’s husbandry, but the seed of the enemy, from whom may you ever be delivered by the prayers of the shepherd, that most faithful and gentle shepherd who presides over you. I therefore exhort you in the Lord to receive with all tenderness those that repent and return to the unity of the Church, that through your kindness and forbearance they may recover\footnote{2 Tim. ii. 26.} themselves out of the snare of the devil, and becoming worthy of Jesus Christ, may obtain eternal salvation in the kingdom of Christ. Brethren, be not deceived. If any man follows him that separates from the truth, he shall not inherit the kingdom of God; and if any man does not stand aloof from the preacher of falsehood, he shall be condemned to hell. For it is obligatory neither to separate from the godly, nor to associate with the ungodly. If any one walks according to a strange\footnote{i.e., heretical.} opinion, he is not of Christ, nor a partaker of His passion; but is a fox,\footnote{Comp.  Cant. ii. 15.} a destroyer of the vineyard of Christ. Have no fellowship\footnote{Comp. 1 Cor. v. 11.} with such a man, lest ye perish along with him, even should he be thy father, thy son, thy brother, or a member of thy family. For says [the Scripture], “Thine eye shall not spare him.”\footnote{Deut. xiii. 6, 18.} You ought therefore to “hate those that hate God, and to waste away [with grief] on account of His enemies.”\footnote{Ps. cxix. 21.} I do not mean that you should beat them or persecute them, as do the Gentiles “that know not the Lord and God;”\footnote{1 Thess. iv. 5.} but that you should regard them as your enemies, and separate yourselves from them, while yet you admonish them, and exhort them to repentance, if it may be they will hear, if it may be they will submit themselves. For our God is a lover of mankind, and “will have all men to be saved, and to come to the knowledge of the truth.”\footnote{1 Tim. ii. 4.} Wherefore “He makes His sun to rise upon the evil and on the good, and sendeth rain on the just and on the unjust;”\footnote{Matt. v. 45.} of whose kindness the Lord, wishing us also to be imitators, says, “Be ye perfect, even as also your Father that is in heaven is perfect.”\footnote{Matt. v. 48.}

\xsection{Chapter IV.—Have but one Eucharist, etc.}

\textsc{Shorter.}~Take ye heed, then, to have but one Eucharist. For there is one flesh of our Lord Jesus Christ, and one cup to [show forth\footnote{Literally, “into.”}] the unity of His blood; one altar; as there is one bishop, along with the presbytery and deacons, my fellow-servants: that so, whatsoever ye do, ye may do it according to [the will of] God.

\textsc{Longer.}~I have confidence of you in the Lord, that ye will be of no other mind. Wherefore I write boldly to your love, which is worthy of God, and exhort you to have but one faith, and one [kind of] preaching, and one Eucharist. For there is one flesh of the Lord Jesus Christ; and His blood which was shed for us is one; one loaf also is broken to all [the communicants], and one cup is distributed among them all: there is but one altar for the whole Church, and one bishop, with the presbytery and deacons, my fellow-servants. Since, also, there is but one unbegotten Being, God, even the Father; and one only-begotten Son, God, the Word and man; and one Comforter, the Spirit of truth; and also one preaching, and one faith, and one baptism;\footnote{Eph. iv. 5.} and one Church which the holy apostles established from one end of the earth to the other by the blood of Christ, and by their own sweat and toil; it behoves you also, therefore, as “a peculiar people, and a holy nation,”\footnote{Tit. ii. 14; 1 Pet. ii. 9} to perform all things with harmony in Christ. Wives, be ye subject to your husbands in the fear of God;\footnote{Eph. v. 22.} and ye virgins, to Christ in purity, not counting marriage an abomination, but desiring that which is better, not for the reproach of wedlock, but for the sake of meditating on the law. Children, obey your parents, and have an affection for them, as workers together with God for your birth [into the world]. Servants, be subject to your masters in God, that ye may be the freed-men of Christ.\footnote{1 Cor. vii. 22.} Husbands, love your wives, as fellow-servants of God, as your own body, as the partners of your life, and your co-adjutors in the procreation of children. Virgins, have Christ alone before your eyes, and His Father in your prayers, being enlightened by the Spirit. May I have pleasure in your purity, as that of Elijah, or as of Joshua the son of Nun, as of Melchizedek, or as of Elisha, as of Jeremiah, or as of John the Baptist, as of the beloved disciple, as of Timothy, as of Titus, as of Evodius, as of Clement, who departed this life in [perfect] chastity,\footnote{There was a prevalent opinion among the ancient Christian writers, that all these holy men lived a life of [chaste] celibacy.} Not, however, that I blame the other blessed [saints] because they entered into the married state, of which I have just spoken.\footnote{Or, “it is not because, etc., that I have mentioned these.”} For I pray that, being found worthy of God, I may be found at their feet in the kingdom, as at the feet of Abraham, and Isaac, and Jacob; as of Joseph, and Isaiah, and the rest of the prophets; as of Peter, and Paul, and the rest of the apostles, that were married men. For they entered into these marriages not for the sake of appetite, but out of regard for the propagation of mankind. Fathers, “bring up your children in the nurture and admonition of the Lord;”\footnote{Eph. vi. 4.} and teach them the holy Scriptures, and also trades, that they may not indulge in idleness. Now [the Scripture] says, “A righteous father educates [his children] well; his heart shall rejoice in a wise son.”\footnote{Prov. xxiii. 24.} Masters, be gentle towards your servants, as holy Job has taught you;\footnote{Job xxxi. 13, 15.} for there is one nature, and one family of mankind. For “in Christ there is neither bond nor free.”\footnote{Gal. iii. 28.} Let governors be obedient to Cæsar; soldiers to those that command them; deacons to the presbyters, as to high-priests; the presbyters, and deacons, and the rest of the clergy, together with all the people, and the soldiers, and the governors, and Cæsar [himself], to the bishop; the bishop to Christ, even as Christ to the Father. And thus unity is preserved throughout. Let not the widows be wanderers about, nor fond of dainties, nor gadders from house to house; but let them be like Judith, noted for her seriousness; and like Anna, eminent for her sobriety. I do not ordain these things as an apostle: for “who am I, or what is my father’s house,”\footnote{1 Sam. xviii. 18; 2 Sam. vii. 18.} that I should pretend to be equal in honour to them? But as your “fellow-soldier,”\footnote{Phil. ii. 25.} I hold the position of one who [simply] admonishes you.

\xsection{Chapter V.—Pray for me.}

\textsc{Shorter.}~My brethren, I am greatly enlarged in loving you; and rejoicing exceedingly [over you], I seek to secure your safety. Yet it is not I, but Jesus Christ, for whose sake being bound I fear the more, inasmuch as I am not yet perfect. But your prayer to God shall make me perfect, that I may attain to that portion which through mercy has been allotted me, while I flee to the Gospel as to the flesh of Jesus, and to the apostles as to the presbytery of the Church. And let us also love the prophets, because they too have proclaimed the Gospel,\footnote{Literally, “have proclaimed in reference to the Gospel.”} and placed their hope in Him,\footnote{In Christ.} and waited for Him; in whom also believing, they were saved, through union to Jesus Christ, being holy men, worthy of love and admiration, having had witness borne to them by Jesus Christ, and being reckoned along with [us] in the Gospel of the common hope.

\textsc{Longer.}~My brethren, I am greatly enlarged in loving you; and rejoicing exceedingly [over you], I seek to secure your safety. Yet it is not I, but the Lord Jesus through me; for whose sake being bound, I fear the more, for I am not yet perfect. But your prayer to God shall make me perfect, that I may attain that to which I have been called, while I flee to the Gospel as to the flesh of Jesus Christ, and to the apostles as the presbytery of the Church. I do also love the prophets as those who announced Christ, and as being partakers of the same Spirit with the apostles. For as the false prophets and the false apostles drew [to themselves] one and the same wicked, deceitful, and seducing\footnote{Literally, “people-deceiving.”} spirit; so also did the prophets and the apostles receive from God, through Jesus Christ, one and the same Holy Spirit, who is good, and sovereign,\footnote{Comp. Ps. li. 12 (LXX.).} and true, and the Author of [saving] knowledge.\footnote{Literally, “teaching.”} For there is one God of the Old and New Testament, “one Mediator between God and men,” for the creation of both intelligent and sensitive beings, and in order to exercise a beneficial and suitable providence [over them]. There is also one Comforter, who displayed\footnote{Or, “wrought.”} His power in Moses, and the prophets, and apostles. All the saints, therefore, were saved by Christ, hoping in Him, and waiting for Him; and they obtained through Him salvation, being holy ones, worthy of love and admiration, having testimony borne to them by Jesus Christ, in the Gospel of our common hope.

\xsection{Chapter VI.—Do not accept Judaism.}

\textsc{Shorter.}~But if any one preach the Jewish law\footnote{Literally, “Judaism.”} unto you, listen not to him. For it is better to hearken to Christian doctrine from a man who has been circumcised, than to Judaism from one uncircumcised. But if either of such persons do not speak concerning Jesus Christ, they are in my judgment but as monuments and sepulchres of the dead, upon which are written only the names of men. Flee therefore the wicked devices and snares of the prince  of this world, lest at any time being conquered\footnote{Literally, “oppressed.”} by his artifices,\footnote{Or, “will.”} ye grow weak in your love. But be ye all joined together\footnote{Some render, “come together into the same place.”} with an undivided heart. And I thank my God that I have a good conscience in respect to you, and that no one has it in his power to boast, either privately or publicly, that I have burdened\footnote{Apparently by attempting to impose the yoke of Judaism.} any one either in much or in little. And I wish for all among whom I have spoken, that they may not possess that for a testimony against them.

\textsc{Longer.}~If any one preaches the one God of the law and the prophets, but denies Christ to be the Son of God, he is a liar, even as also is his father the devil,\footnote{Comp. John viii 44.} and is a Jew falsely so called, being possessed of\footnote{Literally, “beneath.”} mere carnal circumcision. If any one confesses Christ Jesus the Lord, but denies the God of the law and of the prophets, saying that the Father of Christ is not the Maker of heaven and earth, he has not continued in the truth any more than his father the devil,\footnote{Comp. John viii 44.} and is a disciple of Simon Magus, not of the Holy Spirit. If any one says there is one God, and also confesses Christ Jesus, but thinks the Lord to be a mere man, and not the only-begotten\footnote{Comp. the reading sanctioned by the ancient authorities, John i. 18.} God, and Wisdom, and the Word of God, and deems Him to consist merely of a soul and body, such an one is a serpent, that preaches deceit and error for the destruction of men. And such a man is poor in understanding, even as by name he is an Ebionite.\footnote{From a Hebrew word meaning “poor.”} If any one confesses the truths mentioned,\footnote{Or, “these things.”} but calls lawful wedlock, and the procreation of children, destruction and pollution, or deems certain kinds of food abominable, such an one has the apostate dragon dwelling within him. If any one confesses the Father, and the Son, and the Holy Ghost, and praises the creation, but calls the incarnation merely an appearance, and is ashamed of the passion, such an one has denied the faith, not less than the Jews who killed Christ. If any one confesses these things, and that God the Word did dwell in a human body, being within it as the Word, even as the soul also is in the body, because it was God that inhabited it, and not a human soul, but affirms that unlawful unions are a good thing, and places the highest happiness\footnote{Literally, “the end of happiness.”} in pleasure, as does the man who is falsely called a Nicolaitan, this person can neither be a lover of God, nor a lover of Christ, but is a corrupter of his own flesh, and therefore void of the Holy Spirit, and a stranger to Christ. All such persons are but monuments and sepulchres of the dead, upon which are written only the names of dead men. Flee, therefore, the wicked devices and snares of the spirit which now worketh in the children of this world,\footnote{Comp. Eph. ii. 2.} lest at any time being overcome,\footnote{Literally, “oppressed.”} ye grow weak in your love. But be ye all joined together\footnote{Some render, “come together into the same place.”} with an undivided heart and a willing mind, “being of one accord and of one judgment,”\footnote{Phil. ii. 2.} being always of the same opinion about the same things, both when you are at ease and in danger, both in sorrow and in joy. I thank God, through Jesus Christ, that I have a good conscience in respect to you, and that no one has it in his power to boast, either privately or publicly, that I have burdened any one either in much or in little. And I wish for all among whom I have spoken, that they may not possess that for a testimony against them.

\xsection{Chapter VII.—I have exhorted you to unity.}

\textsc{Shorter.}~For though some would have deceived me according to the flesh, yet the Spirit, as being from God, is not deceived. For it knows both whence it comes and whither it goes,\footnote{John iii. 8.} and detects the secrets [of the heart]. For, when I was among you, I cried, I spoke with a loud voice: Give heed to the bishop, and to the presbytery and deacons. Now, some suspected me of having spoken thus, as knowing beforehand the division caused by some among you.\footnote{Some translate, “as foreseeing the division to arise among you.”} But He is my witness, for whose sake I am in bonds, that I got no intelligence from any man.\footnote{Literally, “did not know from human flesh.”} But the Spirit proclaimed  these words: Do nothing without the bishop; keep your bodies\footnote{Literally, “your flesh.”} as the temples of God;\footnote{Comp. 1 Cor. iii. 16, 1 Cor. vi. 19.} love unity; avoid divisions; be the followers of Jesus Christ, even as He is of His Father.

\textsc{Longer.}~For though some would have deceived me according to the flesh, yet my spirit is not deceived; for I have received it from God. For it knows both whence it comes and whither it goes, and detects the secrets [of the heart]. For when I was among you, I cried, I spoke with a loud voice—the word is not mine, but God’s—Give heed to the bishop, and to the presbytery and deacons. But if ye suspect that I spake thus, as having learned beforehand the division caused by some among you, He is my witness, for whose sake I am in bonds, that I learned nothing of it from the mouth of any man. But the Spirit made an announcement to me, saying as follows: Do nothing without the bishop; keep your bodies\footnote{Literally, “your flesh.”} as the temples of God; love unity; avoid divisions; be ye followers of Paul, and of the rest of the apostles, even as they also were of Christ.

\xsection{Chapter VIII.—The same continued.}

\textsc{Shorter.}~I therefore did what belonged to me, as a man devoted to\footnote{Literally, “prepared for.”} unity. For where there is division and wrath, God doth not dwell. To all them that repent, the Lord grants forgiveness, if they turn in penitence to the unity of God, and to communion with the bishop.\footnote{Literally, “to the assembly of the bishop.”} I trust [as to you] in the grace of Jesus Christ, who shall free you from every bond. And I exhort you to do nothing out of strife, but according to the doctrine of Christ. When I heard some saying, If I do not find it in the ancient\footnote{The meaning here is very doubtful. Some read \textgreek{ἐν τοῖς ἀρχαίοις}, as translated above; others prefer \textgreek{ἐν τοῖς ἀρχείοις}, as in the longer recension.} Scriptures, I will not believe the Gospel; on my saying to them, It is written, they answered me, That remains to be proved. But to me Jesus Christ is in the place of all that is ancient: His cross, and death, and resurrection, and the faith\footnote{i.e., the system of Christian doctrine.} which is by Him, are undefiled monuments of antiquity; by which I desire, through your prayers, to be justified.

\textsc{Longer.}~I therefore did what belonged to me, as a man devoted to unity; adding this also, that where there is diversity of judgment, and wrath, and hatred, God does not dwell. To all them that repent, God grants forgiveness, if they with one consent return to the unity of Christ, and communion with the bishop.\footnote{Literally, “to the assembly of the bishop.”} I trust to the grace of Jesus Christ, that He will free you from every bond of wickedness.\footnote{Comp. Isa. lviii. 6.} I therefore exhort you that ye do nothing out of strife,\footnote{Phil. ii. 3.} but according to the doctrine of Christ. For I have heard some saying, If I do not find the Gospel in the archives, I will not believe it. To such persons I say that my archives are Jesus Christ, to disobey whom is manifest destruction. My authentic archives are His cross, and death, and resurrection, and the faith which bears on these things, by which I desire, through your prayers, to be justified. He who disbelieves the Gospel disbelieves everything along with it. For the archives ought not to be preferred to the Spirit.\footnote{Or, “the archives of the Spirit are not exposed to all.”} “It is hard to kick against the pricks;”\footnote{Acts xxvi. 14.} it is hard to disbelieve Christ; it is hard to reject the preaching of the apostles.

\xsection{Chapter IX.—The Old Testament is good: the New Testament is better.}

\textsc{Shorter.}~The priests\footnote{i.e., the Jewish priests.} indeed are good, but the High Priest is better; to whom the holy of holies has been committed, and who alone has been trusted with the secrets of God. He is the door of the Father, by which enter in Abraham, and Isaac, and Jacob, and the prophets, and the apostles, and the Church. All these have for their object the attaining to the unity of God. But the Gospel possesses something transcendent [above the former dispensation], viz., the appearance of our Lord Jesus Christ, His passion and resurrection. For the beloved prophets announced Him,\footnote{Literally, “proclaimed as to him.”} but the Gospel is the perfection of immortality.\footnote{The meaning is doubtful. Comp. 2 Tim. i. 10.} All these things are good together, if ye believe in love.

\textsc{Longer.}~The priests\footnote{i.e., the Jewish priests.} indeed, and the ministers of the word, are good; but the High Priest is better, to whom the holy of holies has been committed, and who alone has been entrusted with the secrets of God. The ministering powers of God are good. The Comforter is holy, and the Word is holy, the Son of the Father, by whom He made all things, and exercises a providence over them all. This is the Way\footnote{John xiv. 6.} which leads to the Father, the Rock,\footnote{1 Cor. x. 4.} the Defence,\footnote{Literally, “the hedge.”} the Key, the Shepherd,\footnote{John x. 11.} the Sacrifice, the Door\footnote{John x. 9.} of knowledge, through which have entered Abraham, and Isaac, and Jacob, Moses and all the company of the prophets, and these pillars of the world, the apostles, and the spouse of Christ, on whose account He poured out His own blood, as her marriage portion, that He might redeem her. All these things tend towards the unity of the one and only true God. But the Gospel possesses something transcendent [above the former dispensation], viz. the appearing of our Saviour Jesus Christ, His passion, and the resurrection itself. For those things which the prophets announced, saying, “Until He come for whom it is reserved, and He shall be the expectation of the Gentiles,”\footnote{Gen. xlix. 10.} have been fulfilled in the Gospel, [our Lord saying,] “Go ye and teach all nations, baptizing them in the name of the Father, and of the Son, and of the Holy Ghost.”\footnote{Matt. xxviii. 19.} All then are good together, the law, the prophets, the apostles, the whole company [of others] that have believed through them: only if we love one another.

\xsection{Chapter X.—Congratulate the inhabitants of Antioch on the close of the persecution.}

\textsc{Shorter.}~Since, according to your prayers, and the compassion which ye feel in Christ Jesus, it is reported to me that the Church which is at Antioch in Syria possesses peace, it will become you, as a Church of God, to elect a deacon to act as the ambassador of God [for you] to [the brethren there], that he may rejoice along with them when they are met together, and glorify the name [of God]. Blessed is he in Jesus Christ, who shall be deemed worthy of such a ministry; and ye too shall be glorified. And if ye are willing, it is not beyond your power to do this, for the sake\footnote{Literally, “for the name of.”} of God; as also the nearest Churches have sent, in some cases bishops, and in others presbyters and deacons.

\textsc{Longer.}~Since, according to your prayers, and the compassion which ye feel in Christ Jesus, it is reported to me that the Church which is at Antioch in Syria possesses peace, it will become you, as a Church of God, to elect a bishop to act as the ambassador of God [for you] to [the brethren] there, that it may be granted them to meet together, and to glorify the name of God. Blessed is he in Christ Jesus, who shall be deemed worthy of such a ministry; and if ye be zealous [in this matter], ye shall receive glory in Christ. And if ye are willing, it is not altogether beyond your power to do this, for the sake of\footnote{Literally, “for the name of.”} God; as also the nearest Churches have sent, in some cases bishops, and in others presbyters and deacons.

\xsection{Chapter XI.—Thanks and salutation.}

\textsc{Shorter.}~Now, as to Philo the deacon, of Cilicia, a man of reputation, who still ministers to me in the word of God, along with Rheus Agathopus, an elect man, who has followed me from Syria, not regarding\footnote{Literally, “bidding farewell to.”} his life,—these bear witness in your behalf; and I myself give thanks to God for you, that ye have received them, even as the Lord you. But may those that dishonoured them be forgiven through the grace of Jesus Christ! The love of the brethren at Troas salutes you; whence also I write to you by Burrhus, who was sent along with me by the Ephesians and Smyrnæans, to show their respect.\footnote{Or, “for the sake of honour.”} May the Lord Jesus Christ honour them, in whom they hope, in flesh, and soul, and faith, and love, and concord! Fare ye well in Christ Jesus, our common hope.

\textsc{Longer.}~Now, as to Philo the deacon, a man of Cilicia, of high reputation, who still ministers to me in the word of God, along with Gaius and Agathopus, an elect man, who has followed me from Syria, not regarding\footnote{Literally, “bidding farewell to.”} his life,—these also bear testimony in your behalf. And I myself give thanks to God for you, because ye have received them: and the Lord will also receive you. But may those that dishonoured them be forgiven through the grace of Jesus Christ, “who wisheth not the death of the sinner, but his repentance.”\footnote{Comp. Ezek. xviii. 23, 32, Ezek. xxxiii. 11; 2 Pet. iii. 9.} The love of the brethren at Troas salutes you; whence also I write to you by Burrhus,\footnote{The \textsc{ms.} has “Burgus.”} who was sent along with me by the Ephesians and Smyrnæans, to show their respect:\footnote{Or, “for the sake of honour.”} whom the Lord Jesus Christ will requite, in whom they hope, in flesh, and soul, and spirit, and faith, and love, and concord. Fare ye well in the Lord Jesus Christ, our common hope, in the Holy Ghost.

\xchapter{Epistle to the Smyrnæans}

\textsc{Shorter.}~\emph{Ignatius, who is also called Theophorus, to the Church of God the Father, and of the beloved Jesus Christ, which has through mercy obtained every kind of gift, which is filled with faith and love, and is deficient in no gift, most worthy of God, and adorned with holiness:}\footnote{Literally, “holy-bearing.”}\emph{ the Church which is at Smyrna, in Asia, wishes abundance of happiness, through the immaculate Spirit and word of God.}

\textsc{Longer.}~\emph{Ignatius, who is also called Theophorus, to the Church of God the most high Father, and His beloved Son Jesus Christ, which has through mercy obtained every kind of gift, which is filled with faith and love, and is deficient in no gift, most worthy of God, and adorned with holiness:}\footnote{Literally, “holy-bearing.”}\emph{ the Church which is at Smyrna, in Asia, wishes abundance of happiness, through the immaculate Spirit and word of God.}

\xsection{Chapter I.—Thanks to God for your faith.}

\textsc{Shorter.}~\textsc{I Glorify} God, even Jesus Christ, who has given you such wisdom. For I have observed that ye are perfected in an immoveable faith, as if ye were nailed to the cross of our Lord Jesus Christ, both in the flesh and in the spirit, and are established in love through the blood of Christ, being fully persuaded with respect to our Lord, that He was truly of the seed of David according to the flesh,\footnote{Rom. i. 3.} and the Son of God according to the will and power\footnote{Theodoret, in quoting this passage, reads, “the Godhead and power.”} of God; that He was truly born of a virgin, was baptized by John, in order that all righteousness might be fulfilled\footnote{Matt. iii. 15.} by Him; and was truly, under Pontius Pilate and Herod the tetrarch, nailed [to the cross] for us in His flesh. Of this fruit\footnote{i.e., the cross, “fruit” being put for \emph{Christ on the tree}.} we are by His divinely-blessed passion, that He might set up a standard\footnote{Isa. v. 26, Isa. xlix. 22.} for all ages, through His resurrection, to all His holy and faithful [followers], whether among Jews or Gentiles, in the one body of His Church.

\textsc{Longer.}~\textsc{I Glorify} the God and Father of our Lord Jesus Christ, who by Him has given you such wisdom. For I have observed that ye are perfected in an immoveable faith, as if ye were nailed to the cross of our Lord Jesus Christ, both in the flesh and in the spirit, and are established in love through the blood of Christ, being fully persuaded, in very truth, with respect to our Lord Jesus Christ, that He was the Son of God, “the first-born of every creature,”\footnote{Col. i. 15.} God the Word, the only-begotten Son, and was of the seed of David according to the flesh,\footnote{Rom. i. 3.} by the Virgin Mary; was baptized by John, that all righteousness might be fulfilled\footnote{Matt. iii. 15.} by Him; that He lived a life of holiness without sin, and was truly, under Pontius Pilate and Herod the tetrarch, nailed [to the cross] for us in His flesh. From whom we also derive our being,\footnote{Literally, “we are.”} from His divinely-blessed passion, that He might set up a standard for the ages, through His resurrection, to all His holy and faithful [followers], whether among Jews or Gentiles, in the one body of His Church.

\xsection{Chapter II.—Christ’s true passion.}

\textsc{Shorter.}~Now, He suffered all these things for our sakes, that we might be saved. And He suffered truly, even as also He truly raised up Himself, not, as certain unbelievers maintain, that He only seemed to suffer, as they themselves only seem to be [Christians]. And as they believe, so shall it happen unto them, when they shall be divested of their bodies, and be mere evil spirits.\footnote{Or, “seeing that they are phantasmal and diabolical,” as some render, but the above is preferable.}

\textsc{Longer.}~Now, He suffered all these things for us; and He suffered them really, and not in appearance only, even as also He truly rose again. But not, as some of the unbelievers, who are ashamed of the formation of man, and the cross, and death itself, affirm, that in appearance only, and not in truth, He took a body of the Virgin, and suffered only in appearance, forgetting, as they do, Him who said, “The Word was made flesh;”\footnote{John i. 14.} and again, “Destroy this temple, and in three days I will raise it up;”\footnote{John ii. 19.} and once more, “If I be lifted up from the earth, I will draw all men unto Me.”\footnote{John xii. 32.} The Word therefore did dwell in flesh, for “Wisdom built herself an house.”\footnote{Prov. ix. 1.} The Word raised up again His own temple on the third day, when it had been destroyed by the Jews fighting against Christ. The Word, when His flesh was lifted up, after the manner of the brazen serpent in the wilderness, drew all men to Himself for their eternal salvation.\footnote{Num. xxi. 9; John iii. 14.}

\xsection{Chapter III.—Christ was possessed of a body after His resurrection.}

\textsc{Shorter.}~For I know that after His resurrection also He was still possessed of flesh,\footnote{Literally, “in the flesh.”} and I believe that He is so now. When, for instance, He came to those who were with Peter, He said to them, “Lay hold, handle Me, and see that I am not an incorporeal spirit.”\footnote{Literally, “demon.” According to Jerome, this quotation is from the Gospel of the Nazarenes. Comp. Luke xxiv. 39.} And immediately they touched Him, and believed, being convinced both by His flesh and spirit. For this cause also they despised death, and were found its conquerors.\footnote{Literally, “above death.”} And after his resurrection He did eat and drink with them, as being possessed of flesh, although spiritually He was united to the Father.

\textsc{Longer.}~And I know that He was possessed of a body not only in His being born and crucified, but I also know that He was so after His resurrection, and believe that He is so now. When, for instance, He came to those who were with Peter, He said to them, “Lay hold, handle Me, and see that I am not an incorporeal spirit.”\footnote{Literally, “demon.” According to Jerome, this quotation is from the Gospel of the Nazarenes. Comp. Luke xxiv. 39.} “For a spirit hath not flesh and bones, as ye see Me have.”\footnote{Luke xxiv. 39.} And He says to Thomas, “Reach hither thy finger into the print of the nails, and reach hither thy hand, and thrust it into My side;”\footnote{John xx. 27.} and immediately they believed that He was Christ. Wherefore Thomas also says to Him, “My Lord, and my God.”\footnote{John xx. 28.} And on this account also did they despise death, for it were too little to say, indignities and stripes. Nor was this all; but also after He had shown Himself to them, that He had risen indeed, and not in appearance only, He both ate and drank with them during forty entire days. And thus was He, with the flesh, received up in their sight unto Him that sent Him, being with that same flesh to come again, accompanied by glory and power. For, say the [holy] oracles, “This same Jesus, who is taken up from you into heaven, shall so come, in like manner as ye have seen Him go unto heaven.”\footnote{Acts i. 11.} But if they say that He will come at the end of the world without a body, how shall those “see Him that pierced Him,”\footnote{Rev. i. 7.} and when they recognise Him, “mourn for themselves?”\footnote{Zech. xii. 10.} For incorporeal beings have neither form nor figure, nor the aspect\footnote{Or, “mark.”} of an animal possessed of shape, because their nature is in itself simple.

\xsection{Chapter IV.—Beware of these heretics.}

\textsc{Shorter.}~I give you these instructions, beloved, assured that ye also hold the same opinions [as I do]. But I  guard you beforehand from those beasts in the shape of men, whom you must not only not receive, but, if it be possible, not even meet with; only you must pray to God for them, if by any means they may be brought to repentance, which, however, will be very difficult. Yet Jesus Christ, who is our true life, has the power of [effecting] this. But if these things were done by our Lord only in appearance, then am I also only in appearance bound. And why have I also surrendered myself to death, to fire, to the sword, to the wild beasts? But, [in fact,] he who is near to the sword is near to God; he that is among the wild beasts is in company with God; provided only he be so in the name of Jesus Christ. I undergo all these things that I may suffer together with Him,\footnote{Comp. Rom. viii. 17.} He who became a perfect man inwardly strengthening me.\footnote{Comp. Phil. iv. 13.}

\textsc{Longer.}~I give you these instructions, beloved, assured that ye also hold the same opinions [as I do]. But I guard you beforehand from these beasts in the shape of men, from whom you must not only turn away, but even flee from them. Only you must pray for them, if by any means they may be brought to repentance. For if the Lord were in the body in appearance only, and were crucified in appearance only, then am I also bound in appearance only. And why have I also surrendered myself to death, to fire, to the sword, to the wild beasts? But, [in fact,] I endure all things for Christ, not in appearance only, but in reality, that I may suffer together with Him, while He Himself inwardly strengthens me; for of myself I have no such ability.

\xsection{Chapter V.—Their dangerous errors.}

\textsc{Shorter.}~Some ignorantly\footnote{Or, “foolishly.”} deny Him, or rather have been denied by Him, being the advocates of death rather than of the truth. These persons neither have the prophets persuaded, nor the law of Moses, nor the Gospel even to this day, nor the sufferings we have individually endured. For they think also the same thing regarding us.\footnote{i.e., As they imagine Christ to have suffered only in appearance, so they believe that we suffer in vain.} For what does any one profit me, if he commends me, but blasphemes my Lord, not confessing that He was [truly] possessed of a body?\footnote{Literally, “a flesh-bearer.”} But he who does not acknowledge this, has in fact altogether denied Him, being enveloped in death.\footnote{Literally, “a death-bearer.”} I have not, however, thought good to write the names of such persons, inasmuch as they are unbelievers. Yea, far be it from me to make any mention of them, until they repent and return to [a true belief in] Christ’s passion, which is our resurrection.

\textsc{Longer.}~Some have ignorantly denied Him, and advocate falsehood rather than the truth. These persons neither have the prophecies persuaded, nor the law of Moses, nor the Gospel even to this day, nor the sufferings we have individually endured. For they think also the same thing regarding us. For what does it profit, if any one commends me, but blasphemes my Lord, not owning Him to be God incarnate?\footnote{Literally, “a flesh-bearer.”} He that does not confess this, has in fact altogether denied Him, being enveloped in death. I have not, however, thought good to write the names of such persons, inasmuch as they are unbelievers; and far be it from me to make any mention of them, until they repent.

\xsection{Chapter VI—Unbelievers in the blood of Christ shall be condemned.}

\textsc{Shorter.}~Let no man deceive himself. Both the things which are in heaven, and the glorious angels,\footnote{Literally, “the glory of the angels.”} and rulers, both visible and invisible, if they believe not in the blood of Christ, shall, in  consequence, incur condemnation.\footnote{Literally, “judgment is to them.”} “He that is able to receive it, let him receive it.”\footnote{Matt. xix. 12.} Let not [high] place puff any one up: for that which is worth all is\footnote{Literally, “the whole is.”} faith and love, to which nothing is to be preferred. But consider those who are of a different opinion with respect to the grace of Christ which has come unto us, how opposed they are to the will of God. They have no regard for love; no care for the widow, or the orphan, or the oppressed; of the bond, or of the free; of the hungry, or of the thirsty.

\textsc{Longer.}~Let no man deceive himself. Unless he believes that Christ Jesus has lived in the flesh, and shall confess His cross and passion, and the blood which He shed for the salvation of the world, he shall not obtain eternal life, whether he be a king, or a priest, or a ruler, or a private person, a master or a servant, a man or a woman. “He that is able to receive it, let him receive it.”\footnote{Matt. xix. 12.} Let no man’s place, or dignity, or riches, puff him up; and let no man’s low condition or poverty abase him. For the chief points are faith towards God, hope towards Christ, the enjoyment of those good things for which we look, and love towards God and our neighbour. For, “Thou shall love the Lord thy God with all thy heart, and thy neighbour as thyself.”\footnote{Deut. vi. 5.} And the Lord says, “This is life eternal, to know the only true God, and Jesus Christ whom He has sent.”\footnote{John xvii. 31.} And again, “A new commandment give I unto you, that ye love one another. On these two commandments hang all the law and the prophets.”\footnote{John xiii. 34; Matt. xxii. 40.} Do ye, therefore, notice those who preach other doctrines, how they affirm that the Father of Christ cannot be known, and how they exhibit enmity and deceit in their dealings with one another. They have no regard for love; they despise the good things we expect hereafter; they regard present things as if they were durable; they ridicule him that is in affliction; they laugh at him that is in bonds.

\xsection{Chapter VII.—Let us stand aloof from such heretics.}

\textsc{Shorter.}~They abstain from the Eucharist and from prayer,\footnote{Theodoret, in quoting this passage, reads \textgreek{προσφοράς}, “offering.”} because they confess not the Eucharist to be the flesh of our Saviour Jesus Christ, which suffered for our sins, and which the Father, of His goodness, raised up again. Those, therefore, who speak against this gift of God, incur death\footnote{Literally, “die disputing.”} in the midst of their disputes. But it were better for them to treat it with respect,\footnote{Literally, “to love.” Some think there is a reference to the \emph{agapæ}, or \emph{love-feasts}.} that they also might rise again. It is fitting, therefore, that ye should keep aloof from such persons, and not to speak of\footnote{The reading is \textgreek{περί} in the one case, and \textgreek{μετά} in the other, though the latter meaning seems preferable. Most of the \textsc{mss.} of the longer recension read \textgreek{περί}, as in the shorter.} them either in private or in public, but to give heed to the prophets, and above all, to the Gospel, in which the passion [of Christ] has been revealed to us, and the resurrection has been fully proved.\footnote{Literally, “perfected.”} But avoid all divisions, as the beginning of evils.

\textsc{Longer.}~They are ashamed of the cross; they mock at the passion; they make a jest of the resurrection. They are the offspring of that spirit who is the author of all evil, who led Adam,\footnote{Literally, “drove Adam out of.”} by means of his wife, to transgress the commandment, who slew Abel by the hands of Cain, who fought against Job, who was the accuser of Joshua\footnote{Zech. iii. 1.} the son of Josedech, who sought to “sift the faith”\footnote{Luke xxii. 31.} of the apostles, who stirred up the multitude of the Jews against the Lord, who also now “worketh in the children of disobedience;\footnote{Eph. ii. 2.} from whom the Lord Jesus Christ will deliver us, who prayed that the faith of the apostles might not fail,\footnote{Luke xxii. 32.} not because He was not able of Himself to preserve it, but because He rejoiced in the pre-eminence of the Father. It is fitting, therefore, that ye should keep aloof from such persons, and neither in private nor in public to talk with\footnote{The reading is \textgreek{περί} in the one case, and \textgreek{μετά} in the other, though the latter meaning seems preferable. Most of the \textsc{mss.} of the longer recension read \textgreek{περί}, as in the shorter.} them; but to give heed to the law, and the prophets, and to those who have preached to you the word of salvation. But flee from all abominable heresies, and those that cause schisms, as the beginning of evils.

\xsection{Chapter VIII.—Let nothing be done without the bishop.}

\textsc{Shorter.}~See that ye all follow the bishop, even as Jesus Christ does the Father, and the presbytery as ye would the apostles; and reverence the deacons, as being the institution\footnote{Or, “command.”} of God. Let no man do anything connected with the Church without the bishop. Let that be deemed a proper\footnote{Or, “firm.”} Eucharist, which is [administered] either  by the bishop, or by one to whom he has entrusted it. Wherever the bishop shall appear, there let the multitude [of the people] also be; even as, wherever Jesus Christ is, there is the Catholic Church. It is not lawful without the bishop either to baptize or to celebrate a love-feast; but whatsoever he shall approve of, that is also pleasing to God, so that everything that is done may be secure and valid.\footnote{Or, “firm.”}

\textsc{Longer.}~See that ye all follow the bishop, even as Christ Jesus does the Father, and the presbytery as ye would the apostles. Do ye also reverence the deacons, as those that carry out [through their office] the appointment of God. Let no man do anything connected with the Church without the bishop. Let that be deemed a proper\footnote{Or, “firm.”} Eucharist, which is [administered] either by the bishop, or by one to whom he has entrusted it. Wherever the bishop shall appear, there let the multitude [of the people] also be; even as where Christ is, there does all the heavenly host stand by, waiting upon Him as the Chief Captain of the Lord’s might, and the Governor of every intelligent nature. It is not lawful without the bishop either to baptize, or to offer, or to present sacrifice, or to celebrate a love-feast.\footnote{Some refer the words to the Lord’s Supper.} But that which seems good to him, is also well-pleasing to God, that everything ye do may be secure and valid.

\xsection{Chapter IX.—Honour the bishop.}

\textsc{Shorter.}~Moreover,\footnote{Or, “finally.”} it is in accordance with reason that we should return to soberness [of conduct], and, while yet we have opportunity, exercise repentance towards God. It is well to reverence\footnote{Literally, “to know.”} both God and the bishop. He who honours the bishop has been honoured by God; he who does anything without the knowledge of the bishop, does [in reality] serve the devil. Let all things, then, abound to you through grace, for ye are worthy. Ye have refreshed me in all things, and Jesus Christ [shall refresh] you. Ye have loved me when absent as well as when present. May God recompense you, for whose sake, while ye endure all things, ye shall attain unto Him.

\textsc{Longer.}~Moreover, it is in accordance with reason that we should return to soberness [of conduct], and, while yet we have opportunity, exercise repentance towards God. For “in Hades there is no one who can confess his sins.”\footnote{Ps. vi. 5.} For “behold the man, and his work is before him.”\footnote{Isa. lxii. 11.} And [the Scripture saith], “My son, honour thou God and the king.”\footnote{Prov. xxiv. 21.} And say I, Honour thou God indeed, as the Author and Lord of all things, but the bishop as the high-priest, who bears the image of God—of God, inasmuch as he is a ruler, and of Christ, in his capacity of a priest. After Him, we must also honour the king. For there is no one superior to God, or even like to Him, among all the beings that exist. Nor is there any one in the Church greater than the bishop, who ministers as a priest to God for the salvation of the whole world. Nor, again, is there any one among rulers to be compared with the king, who secures peace and good order to those over whom he rules. He who honours the bishop shall be honoured by God, even as he that dishonours him shall be punished by God. For if he that rises up against kings is justly held worthy of punishment, inasmuch as he dissolves public order, of how much sorer punishment, suppose ye, shall he be thought worthy,\footnote{Comp. Heb. x. 29.} who presumes to do anything without the bishop, thus both destroying the [Church’s] unity, and throwing its order into confusion? For the priesthood is the very highest point of all good things among men, against which whosoever is mad enough to strive, dishonours not man, but God, and Christ Jesus, the First-born, and the only High Priest, by nature, of the Father. Let all things therefore be done by you with good order in Christ. Let the laity be subject to the deacons; the deacons to the presbyters; the presbyters to the bishop; the bishop to Christ, even as He is to the Father. As ye, brethren, have refreshed me, so will Jesus Christ refresh you. Ye have loved me when absent, as well as when present. God will recompense you, for whose sake ye have shown such kindness towards His prisoner. For even if I am not worthy of it, yet your zeal [to help me] is an admirable\footnote{Or, “great.”} thing. For “he who honours a prophet in the name of a prophet, shall receive a prophet’s reward.”\footnote{Matt. x. 41.} It is manifest also, that he who honours a prisoner of Jesus Christ shall receive the reward of the martyrs.

\xsection{Chapter X.—Acknowledgment of their kindness.}

\textsc{Shorter.}~Ye have done well in receiving Philo and Rheus Agathopus as servants\footnote{Or, “deacons.”} of Christ our God, who have followed me for the sake of God, and who give thanks to the Lord in your behalf, because ye have in every way refreshed them. None of these things shall be lost to you. May my spirit be for you,\footnote{Comp. Epistle of Ignatius to Ephesians, chap. xxi.; to Polycarp, chap. ii. vi.} and my bonds, which ye have not despised or been ashamed of; nor shall Jesus Christ, our perfect hope, be ashamed of you.

\textsc{Longer.}~Ye have done well in receiving Philo, and Gaius, and Agathopus, who, being the servants\footnote{Or, “deacons.”} of Christ, have followed me for the sake of God, and who greatly bless the Lord in your behalf, because ye have in every way refreshed them. None of those things which ye have done to them shall be passed by without being reckoned unto you. “The Lord grant” to you “that ye may find mercy of the Lord in that day!”\footnote{2 Tim. i. 18.} May my spirit be for you,\footnote{Comp. Epistle of Ignatius to Ephesians, chap. xxi.; to Polycarp, chap. ii. vi.} and my bonds, which ye have not despised or been ashamed of. Wherefore, neither shall Jesus Christ, our perfect hope, be ashamed of you.

\xsection{Chapter XI.—Request to them to send a messenger to Antioch.}

\textsc{Shorter.}~Your prayer has reached to the Church which is at Antioch in Syria. Coming from that place bound with chains, most acceptable to God,\footnote{Literally, “most becoming of God.”} I salute all; I who am not worthy to be styled from thence, inasmuch as I am the least of them. Nevertheless, according to the will of God, I have been thought worthy [of this honour], not that I have any sense\footnote{Or, “from any conscience.”} [of having deserved it], but by the grace of God, which I wish may be perfectly given to me, that through your prayers I may attain to God. In order, therefore, that your work may be complete both on earth and in heaven, it is fitting that, for the honour of God, your Church should elect some worthy delegate;\footnote{Literally, “God-ambassador.”} so that he, journeying into Syria, may congratulate them that they are [now] at peace, and are restored to\footnote{Or, “having received.”} their proper greatness, and that their proper constitution\footnote{Literally, “body.”} has been re-established among them. It seems then to me a becoming thing, that you should send some one of your number with an epistle, so that, in company with them, he may rejoice\footnote{Literally, “may glorify with him.”} over the tranquillity which, according to the will of God, they have obtained, and because that, through your prayers, they have now reached the harbour. As persons who are perfect, ye should also aim at\footnote{Or, “think of.”} those things which are perfect. For when ye are desirous to do well, God is also ready to assist you.

\textsc{Longer.}~Your prayers have reached to the Church of Antioch, and it is at peace. Coming from that place bound, I salute all; I who am not worthy to be styled from thence, inasmuch as I am the least of them. Nevertheless, according to the will of God, I have been thought worthy [of this honour], not that I have any sense\footnote{Or, “from any conscience.”} [of having deserved it], but by the grace of God, which I wish may be perfectly given to me, that through your prayers I may attain to God. In order, therefore, that your work may be complete both on earth and in heaven, it is fitting that, for the honour of God, your Church should elect some worthy delegate;\footnote{Literally, “God-ambassador.”} so that he, journeying into Syria, may congratulate them that they are [now] at peace, and are restored to their proper greatness, and that their proper constitution\footnote{Literally, “body.”} has been re-established among them. What appears to me proper to be done is this, that you should send some one of your number with an epistle, so that, in company with them, he may rejoice over the tranquillity which, according to the will of God, they have obtained, and because that, through your prayers, I have secured Christ as a safe harbour. As persons who are perfect, ye should also aim at\footnote{Or, “think of.”} those things which are perfect. For when ye are desirous to do well, God is also ready to assist you.

\xsection{Chapter XII.—Salutations.}

\textsc{Shorter.}~The love of the brethren at Troas salutes you; whence also I write to  you by Burrhus, whom ye sent with me, together with the Ephesians, your brethren, and who has in all things refreshed me. And I would that all may imitate him, as being a pattern of a minister\footnote{Or, “the ministry.”} of God. Grace will reward him in all things. I salute your most worthy\footnote{Literally, “worthy of God.”} bishop, and your very venerable\footnote{Literally, “most becoming of God.”} presbytery, and your deacons, my fellow-servants, and all of you individually, as well as generally, in the name of Jesus Christ, and in His flesh and blood, in His passion and resurrection, both corporeal and spiritual, in union with God and you.\footnote{Literally, “in the union of God and of you.”} Grace, mercy, peace, and patience, be with you for evermore!

\textsc{Longer.}~The love of your brethren at Troas salutes you; whence also I write to you by Burgus, whom ye sent with me, together with the Ephesians, your brethren, and who has in all things refreshed me. And I would that all may imitate him, as being a pattern of a minister of God. The grace of the Lord will reward him in all things. I salute your most worthy bishop Polycarp, and your venerable presbytery, and your Christ-bearing deacons, my fellow-servants, and all of you individually, as well as generally, in the name of Christ Jesus, and in His flesh and blood, in His passion and resurrection, both corporeal and spiritual, in union with God and you. Grace, mercy, peace, and patience, be with you in Christ for evermore!

\xsection{Chapter XIII.—Conclusion.}

\textsc{Shorter.}~I salute the families of my brethren, with their wives and children, and the virgins who are called widows.\footnote{The \emph{deaconesses} seem to have been called \emph{widows}.} Be ye strong, I pray, in the power of the Holy Ghost. Philo, who is with me, greets you. I salute the house of Tavias, and pray that it may be confirmed in faith and love, both corporeal and spiritual. I salute Alce, my well-beloved,\footnote{Literally, “the name desired of me.”} and the incomparable Daphnus, and Eutecnus, and all by name. Fare ye well in the grace of God.

\textsc{Longer.}~I salute the families of my brethren, with their wives and children, and those that are ever virgins, and the widows. Be ye strong, I pray, in the power of the Holy Ghost. Philo, my fellow-servant, who is with me, greets you. I salute the house of Tavias, and pray that it may be confirmed in faith and love, both corporeal and spiritual. I salute Alce, my well-beloved,\footnote{Literally, “the name desired of me.”} and the incomparable Daphnus, and Eutecnus, and all by name. Fare ye well in the grace of God, and of our Lord Jesus Christ, being filled with the Holy Spirit, and divine and sacred wisdom.

\xchapter{Epistle to Polycarp}

\textsc{Shorter.}~\emph{Ignatius, who is also called Theophorus, to Polycarp, Bishop of the Church of the Smyrnæans, or rather, who has, as his own bishop, God the Father, and the Lord Jesus Christ: [wishes] abundance of happiness.}

\textsc{Longer.}~\emph{Ignatius, bishop of Antioch, and a witness for Jesus Christ, to Polycarp, Bishop of the Church of the Smyrnæans, or rather, who has, as his own bishop, God the Father, and Jesus Christ: [wishes] abundance of happiness.}

\xsection{Chapter I.—Commendation and exhortation.}

\textsc{Shorter.}~\textsc{Having} obtained good proof that thy mind is fixed in God as upon an immoveable rock, I loudly glorify [His name] that I have been thought worthy [to behold] thy blameless face,\footnote{i.e., to make personal acquaintance with one esteemed so highly.} which may I ever enjoy in God! I entreat thee, by the grace with which thou art clothed, to press forward in thy course, and to exhort all that they may be saved. Maintain thy position with all care, both in the flesh and spirit. Have a regard to preserve unity, than which nothing is better. Bear with all, even as the Lord does with thee. Support\footnote{Or, “tolerate.”} all in love, as also thou doest. Give thyself to prayer without ceasing.\footnote{Comp. 1 Thess. v. 17.} Implore additional understanding to what thou already hast. Be watchful, possessing a sleepless spirit. Speak to every man separately, as God enables thee.\footnote{Some read, “according to thy practice.”} Bear the infirmities of all, as being a perfect athlete [in the Christian life]: where the labour is great, the gain is all the more.

\textsc{Longer.}~\textsc{Having} obtained good proof that thy mind is fixed in God as upon an immoveable rock, I loudly glorify [His name] that I have been thought worthy to behold thy blameless face,\footnote{i.e., to make personal acquaintance with one esteemed so highly.} which may I ever enjoy in God! I entreat thee, by the grace with which thou art clothed, to press forward in thy course, and to exhort all that they may be saved. Maintain thy position with all care, both in the flesh and spirit. Have a regard to preserve unity, than which nothing is better. Bear with all even as the Lord does with thee. Support\footnote{Or, “tolerate.”} all in love, as also thou doest. Give thyself to prayer without ceasing.\footnote{Comp. 1 Thess. v. 17.} Implore additional understanding to what thou already hast. Be watchful, possessing a sleepless spirit. Speak to every man separately, as God enables thee.\footnote{Some read, “according to thy practice.”} Bear the infirmities of all, as being a perfect athlete [in the Christian life], even as does the Lord of all. For says [the Scripture], “He Himself took our infirmities, and bare our sicknesses.”\footnote{Matt. viii. 17.} Where the labour is great, the gain is all the more.

\xsection{Chapter II.—Exhortations.}

\textsc{Shorter.}~If thou lovest the good disciples, no thanks are due to thee on that account; but rather seek by meekness to subdue the more troublesome. Every kind of wound is not healed with the same plaster. Mitigate violent  attacks [of disease] by gentle applications.\footnote{Literally, “paroxysms by embrocations.”} Be in all things “wise as a serpent, and harmless as a dove.”\footnote{Matt. x. 16.} For this purpose thou art composed of both flesh and spirit, that thou mayest deal tenderly\footnote{Literally, “flatter.”} with those [evils] that present themselves visibly before thee. And as respects those that are not seen,\footnote{Some refer this to the mysteries of God and others to things yet future.} pray that [God] would reveal them unto thee, in order that thou mayest be wanting in nothing, but mayest abound in every gift. The times call for thee, as pilots do for the winds, and as one tossed with tempest seeks for the haven, so that both thou [and those under thy care] may attain to God. Be sober as an athlete of God: the prize set before thee is immortality and eternal life, of which thou art also persuaded. In all things may my soul be for thine,\footnote{Comp. Epistle of Ignatius to the Ephesians, chap. xxi., etc.} and my bonds also, which thou hast loved.

\textsc{Longer.}~If thou lovest the good disciples, no thanks are due to thee on that account; but rather seek by meekness to subdue the more troublesome. Every kind of wound is not healed with the same plaster. Mitigate violent attacks [of disease] by gentle applications.\footnote{Literally, “paroxysms by embrocations.”} Be in all things “wise as a serpent, and harmless always as a dove.”\footnote{Matt. x. 16.} For this purpose thou art composed of both soul and body, art both fleshly and spiritual, that thou mayest correct those [evils] that present themselves visibly before thee; and as respects those that are not seen, mayest pray that these should be revealed to thee, so that thou mayest be wanting in nothing, but mayest abound in every gift. The times call upon thee to pray. For as the wind aids the pilot of a ship, and as havens are advantageous for safety to a tempest-tossed vessel, so is also prayer to thee, in order that thou mayest attain to God. Be sober as an athlete of God, whose will is immortality and eternal life; of which thou art also persuaded. In all things may my soul be for thine,\footnote{Comp. Epistle of Ignatius to the Ephesians, chap. xxi., etc.} and my bonds also, which thou hast loved.

\xsection{Chapter III.—Exhortations.}

\textsc{Shorter.}~Let not those who seem worthy of credit, but teach strange doctrines,\footnote{Comp. 1 Tim. i. 3, 1 Tim. vi. 3.} fill thee with apprehension. Stand firm, as does an anvil which is beaten. It is the part of a noble\footnote{Literally, “great.”} athlete to be wounded, and yet to conquer. And especially, we ought to bear all things for the sake of God, that He also may bear with us. Be ever becoming more zealous than what thou art. Weigh carefully the times. Look for Him who is above all time, eternal and invisible, yet who became visible for our sakes; impalpable and impassible, yet who became passible on our account; and who in every kind of way suffered for our sakes.

\textsc{Longer.}~Let not those who seem worthy of credit, but teach strange doctrines,\footnote{Comp. 1 Tim. i. 3, 1 Tim. vi. 3.} fill thee with apprehension. Stand firm, as does an anvil which is beaten. It is the part of a noble\footnote{Literally, “great.”} athlete to be wounded, and yet to conquer. And especially we ought to bear all things for the sake of God, that He also may bear with us, and bring us into His kingdom. Add more and more to thy diligence; run thy race with increasing energy; weigh carefully the times. Whilst thou art here, be a conqueror; for here is the course, and there are the crowns. Look for Christ, the Son of God; who was before time, yet appeared in time; who was invisible by nature, yet visible in the flesh; who was impalpable, and could not be touched, as being without a body, but for our sakes became such, might be touched and handled in the body; who was impassible as God, but became passible for our sakes as man; and who in every kind of way suffered for our sakes.

\xsection{Chapter IV.—Exhortations.}

\textsc{Shorter.}~Let not widows be neglected. Be thou, after the Lord, their protector\footnote{The word in the original (\textgreek{φροντιστής}) denotes one who \emph{thinks} or \emph{cares} for another.} and friend. Let nothing be done without thy consent; neither do thou anything without the approval of God, which indeed thou dost not, inasmuch as thou art stedfast. Let your assembling together be of frequent\footnote{Some refer the words to more \emph{frequent} meetings, and others to these meetings being more numerous; no comparison is necessarily implied.} occurrence: seek after all by name.\footnote{i.e., so as to bring them out to the public assembly.} Do not despise either male or female slaves, yet neither let them be puffed up with conceit, but rather let them  submit themselves\footnote{Or, “act the part of slaves.”} the more, for the glory of God, that they may obtain from God a better liberty. Let them not long to be set free [from slavery] at the public expense, that they be not found slaves to their own desires.

\textsc{Longer.}~Let not the widows be neglected. Be thou, after the Lord, their protector and friend. Let nothing be done without thy consent; neither do thou anything without the approval of God, which indeed thou doest not. Be thou stedfast. Let your assembling together be of frequent\footnote{Some refer the words to more \emph{frequent} meetings, and others to these meetings being more numerous; no comparison is necessarily implied.} occurrence: seek after all by name.\footnote{i.e., so as to bring them out to the public assembly.} Do not despise either male or female slaves, yet neither let them be puffed up with conceit, but rather let them submit themselves\footnote{Or, “act the part of slaves.”} the more, for the glory of God, that they may obtain from God a better liberty. Let them not wish to be set free [from slavery] at the public expense, that they be not found slaves to their own desires.

\xsection{Chapter V.—The duties of husbands and wives.}

\textsc{Shorter.}~Flee evil arts; but all the more discourse in public regarding them.\footnote{Some insert \textgreek{μή}, and render, “rather do not even speak of them.”} Speak to my sisters, that they love the Lord, and be satisfied with their husbands both in the flesh and spirit. In like manner also, exhort my brethren, in the name of Jesus Christ, that they love their wives, even as the Lord the Church.\footnote{Eph. v. 25.} If any one can continue in a state of purity,\footnote{i.e., in celibacy.} to the honour of Him who is Lord of the flesh,\footnote{Some render, “to the honour of the flesh of the Lord,” as in the longer recension.} let him so remain without boasting. If he begins to boast, he is undone; and if he reckon himself greater than the bishop, he is ruined. But it becomes both men and women who marry, to form their union with the approval of the bishop, that their marriage may be according to God, and not after their own lust. Let all things be done to the honour of God.\footnote{Comp. 1 Cor. x. 31.}

\textsc{Longer.}~Flee evil arts; but all the more discourse in public regarding them. Speak to my sisters, that they love the Lord, and be satisfied with their husbands both in the flesh and spirit. In like manner also, exhort my brethren, in the name of Jesus Christ, that they love their wives, even as the Lord the Church. If any one can continue in a state of purity,\footnote{i.e., in celibacy.} to the honour of the flesh of the Lord, let him so remain without boasting. If he shall boast, he is undone; and if he seeks to be more prominent\footnote{Literally, “if he be known beyond the bishop.”} than the bishop, he is ruined. But it becomes both men and women who marry, to form their union with the approval of the bishop, that their marriage may be according to the Lord, and not after their own lust. Let all things be done to the honour of God.\footnote{Comp. 1 Cor. x. 31.}

\xsection{Chapter VI.—The duties of the Christian flock.}

\textsc{Shorter.}~Give ye\footnote{As this Epistle, though sent to the bishop, was meant to be read to the people, Ignatius here directly addresses them.} heed to the bishop, that God also may give heed to you. My soul be for theirs\footnote{Comp. chap. ii. etc.} that are submissive to the bishop, to the presbyters, and to the deacons, and may my portion be along with them in God! Labour together with one another; strive in company together; run together; suffer together; sleep together; and awake together, as the stewards, and associates,\footnote{Or, “assessors.”} and servants of God. Please ye Him under whom ye fight, and from whom ye receive your wages. Let none of you be found a deserter. Let your baptism endure as your arms; your faith as your helmet; your love as your spear; your patience as a complete panoply. Let your works be the charge\footnote{A military reference, simply implying the idea of faithful effort leading to future reward.} assigned to you, that ye may receive a worthy recompense. Be long-suffering, therefore, with one another, in meekness, as God is towards you. May I have joy of you for ever!\footnote{Comp. Ignatius’ Epistle to the Ephesians, chap. ii.}

\textsc{Longer.}~Give ye\footnote{As this Epistle, though sent to the bishop, was meant to be read to the people, Ignatius here directly addresses them.} heed to the bishop, that God also may give heed to you. My soul be for theirs\footnote{Comp. chap. ii. etc.} that are submissive to the bishop, to the presbytery, and to the deacons: may I have my portion with them from God! Labour together with one another; strive in company together; run together; suffer together; sleep together; and awake together, as the stewards, and associates,\footnote{Or, “assessors.”} and servants of God. Please ye Him under whom ye fight, and from whom ye shall receive your wages. Let none of you be found a deserter. Let your baptism endure as your arms; your faith as your helmet; your love as your spear; your patience as a complete panoply. Let your works be the charge assigned to you, that you may obtain for them a most worthy\footnote{Literally, “worthy of God.”} recompense. Be long-suffering, therefore, with one another, in meekness, and God shall be so with you. May I have joy of you for ever!\footnote{Comp. Ignatius’ Epistle to the Ephesians, chap. ii.}

\xsection{Chapter VII.—Request that Polycarp would send a messenger to Antioch.}

\textsc{Shorter.}~Seeing that the Church which is at Antioch in Syria is, as report has informed me, at peace, through your prayers, I also am the more encouraged, resting without anxiety in God,\footnote{Literally, “in freedom from care of God.”} if indeed by means of suffering I may attain to God, so that, through your prayers, I may be found a disciple [of Christ].\footnote{Some read, “in the resurrection.”} It is fitting, O Polycarp, most blessed in God, to assemble a very solemn\footnote{Literally, “most befitting God.”} council, and to elect one whom you greatly love, and know to be a man of activity, who may be designated the messenger of God;\footnote{Literally, “God-runner.”} and to bestow on him this honour that he may go into Syria, and glorify your ever active love to the praise of Christ. A Christian has not power over himself, but must always be ready for\footnote{Literally, “at leisure for.”} the service of God. Now, this work is both God’s and yours, when ye shall have completed it to His glory.\footnote{Literally, “to Him.”} For I trust that, through grace, ye are prepared for every good work pertaining to God. Knowing, therefore, your energetic love of the truth, I have exhorted you by this brief Epistle.

\textsc{Longer.}~Seeing that the Church which is at Antioch in Syria is, as report has informed me, at peace, through your prayers, I also am the more encouraged, resting without anxiety in God,\footnote{Literally, “in freedom from care of God.”} if indeed by means of suffering I may attain to God, so that, through your prayers, I may be found a disciple [of Christ]. It is fitting, O Polycarp, most blessed in God, to assemble a very solemn\footnote{Literally, “most befitting God.”} council, and to elect one whom you greatly love, and know to be a man of activity, who may be designated the messenger of God;\footnote{Literally, “God-runner.”} and to bestow on him the honour of going into Syria, so that, going into Syria, he may glorify your ever active love to the praise of God. A Christian has not power over himself, but must always be ready for\footnote{Literally, “at leisure for.”} the service of God. Now, this work is both God’s and yours, when ye shall have completed it. For I trust that, through grace, ye are prepared for every good work pertaining to God. Knowing your energetic love of the truth, I have exhorted you by this brief Epistle.

\xsection{Chapter VIII.—Let other churches also send to Antioch.}

\textsc{Shorter.}~Inasmuch as I have not been able to write to all the Churches, because I must suddenly sail from Troas to Neapolis, as the will\footnote{Some suppose the reference to be to the soldiers, or perhaps to God Himself.} [of the emperor] enjoins, [I beg that] thou, as being acquainted with the purpose\footnote{Or, “as possessed of the judgment.”} of God, wilt write to the adjacent Churches, that they also may act in like manner, such as are able to do so sending messengers,\footnote{Literally, “men on foot.”} and the others transmitting letters through those persons who are sent by thee, that thou\footnote{Some have the plural “ye” here.} mayest be glorified by a work\footnote{Literally, “an eternal work.”} which shall be remembered for ever, as indeed thou art worthy to be. I salute all by name, and in particular the wife of Epitropus, with all her house and children. I salute Attalus, my beloved. I salute him who shall be deemed worthy to go [from you] into Syria. Grace shall be with him for ever, and with Polycarp that sends him. I pray for your happiness for ever in our God, Jesus Christ, by whom continue ye in the unity and under the protection of God,\footnote{Some propose to read, “and of the bishop.”} I salute Alce, my dearly beloved.\footnote{Literally, “name desired by me.”} Fare ye well in the Lord.

\textsc{Longer.}~Inasmuch, therefore, as I have not been able to write to all Churches, because I must suddenly sail from Troas to Neapolis, as the will\footnote{Some suppose the reference to be to the soldiers, or perhaps to God Himself.} [of the emperor] enjoins, [I beg that] thou, as being acquainted with the purpose\footnote{Or, “as possessed of the judgment.”} of God, wilt write to the adjacent Churches, that they also may act in like manner, such as are able to do so sending messenger, and the others transmitting letters through those persons who are sent by thee, that thou mayest be glorified by a work\footnote{Literally, “an eternal work.”} which shall be remembered for ever, as indeed thou art worthy to be. I salute all by name, and in particular the wife of Epitropus, with all her house and children. I salute Attalus, my beloved. I salute him who shall be deemed worthy to go [from you] into Syria. Grace shall be with him for ever, and with Polycarp that sends him. I pray for your happiness for ever in our God, Jesus Christ, by whom continue ye in the unity and under the protection of God. I salute Alce, my dearly beloved.\footnote{Literally, “name desired by me.”} Amen. Grace [be with you]. Fare ye well in the Lord.
